---
title: "fertilisation_curves_2026"
author: "Diego Moreno"
date: "2026-02-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

load libraries 

```{r}
library(tidyverse)
library(brms)
library(tidybayes)
library(ggdist)
library(scales)
library(cmdstanr)
```

load database

```{r}
db <- read.delim("../Data/fertilisation_curves_publication_2026.txt")
glimpse(db)
```
Helper function for plotting concentration
Function to convert the +1 in conc_log

```{r}

# helper (only needed for plotting on linear concentration)
to_conc <- function(z) (10^z - 1)
```

MCMC settings 

```{r}
# MCMC settings
chains  <- 4
iter    <- 4000
warmup  <- 2000
cores   <- 4
seed    <- 123
```

Prepare *D. dianthus* data

```{r}
df0_dd <- db %>%
  filter(species == "D. dianthus") %>%
  transmute(
    cross,
    y   = as.numeric(fertilised),
    n   = as.numeric(particles),
    x_sc = as.numeric(conc_log)   # log10(conc + 1) already in db
  ) %>%
  filter(!is.na(y), !is.na(n), n > 0, is.finite(x_sc))

if (nrow(df0_dd) == 0) stop("No data found for D. dianthus")

# summary(df0_dd$x_sc)
# range(df0_dd$x_sc)
# df0_dd %>% count(cross)

```
dentify *D. dianthus* crosses that never fertilised

```{r}
never_fert_crosses_dd <- df0_dd %>%
  group_by(cross) %>%
  summarise(total_fert = sum(y), .groups = "drop") %>%
  filter(total_fert == 0) %>%
  pull(cross)

never_fert_crosses_dd

df_model_dd <- df0_dd %>%
  filter(!cross %in% never_fert_crosses_dd)

```

Fit *D. dianthus* hierarchical binomial model

```{r}
priors_dd <- c(
  prior(normal(0, 5), class = "Intercept"),
  prior(normal(0, 3), class = "b"),
  prior(exponential(1), class = "sd", group = "cross"),
  prior(lkj(2), class = "cor")
)

fit_dd <- brm(
  y | trials(n) ~ 1 + x_sc + (1 + x_sc | cross),
  data    = df_model_dd,
  family  = binomial(link = "logit"),
  prior   = priors_dd,
  chains  = chains, iter = iter, warmup = warmup, cores = cores,
  control = list(adapt_delta = 0.97, max_treedepth = 12),
  seed    = seed,
  backend = "cmdstanr"
)

```

Check *D. dianthus* model diagnostics

```{r}
summary(fit_dd)
max(rhat(fit_dd), na.rm = TRUE)
min(neff_ratio(fit_dd), na.rm = TRUE)

plot(fit_dd)

pp_check(fit_dd)
pp_check(fit_dd, type = "bars", ndraws = 50)
```

Classify *D. dianthus* crosses at the benchmark concentration

```{r}

c_bench_dd  <- 1e4
target_prob <- 0.50
p_cut       <- 0.80

newdata_bench_dd <- tibble(
  cross = unique(df_model_dd$cross),
  x_sc  = log10(c_bench_dd + 1),
  n     = 1L
)

bench_draws_dd <- add_epred_draws(
  fit_dd,
  newdata = newdata_bench_dd,
  re_formula = NULL
)

prob_col_dd <- if (".epred" %in% names(bench_draws_dd)) ".epred" else ".value"

bench_summary_dd <- bench_draws_dd %>%
  group_by(cross) %>%
  summarise(
    p_reach_target = mean(!!sym(prob_col_dd) >= target_prob),
    prob_median    = median(!!sym(prob_col_dd)),
    .groups = "drop"
  )

cross_groups_dd <- bench_summary_dd %>%
  mutate(
    line_group = if_else(
      p_reach_target >= p_cut,
      "D. dianthus – High",
      "D. dianthus – Low"
    ),
    species = "D. dianthus"
  )

high_crosses_dd <- cross_groups_dd %>%
  filter(line_group == "D. dianthus – High") %>%
  pull(cross)

low_crosses_dd <- cross_groups_dd %>%
  filter(line_group == "D. dianthus – Low") %>%
  pull(cross)
```

Generate group-average curves for *D. dianthus*

```{r}
n_grid <- 120

make_group_curve <- function(fit, df_model, crosses, species_name, group_name, n_grid = 120) {
  if (length(crosses) == 0) return(tibble())

  r <- df_model %>%
    filter(cross %in% crosses) %>%
    summarise(min_x = min(x_sc), max_x = max(x_sc))

  grid <- tibble(
    x_sc = seq(r$min_x, r$max_x, length.out = n_grid),
    n    = 1L
  )

  newdata <- expand_grid(cross = crosses, grid)

  draws <- add_epred_draws(
    fit,
    newdata = newdata,
    re_formula = NULL
  )

  prob_col <- if (".epred" %in% names(draws)) ".epred" else ".value"

  draws %>%
    rename(prob = !!sym(prob_col)) %>%
    group_by(.draw, x_sc) %>%
    summarise(prob = mean(prob), .groups = "drop") %>%
    group_by(x_sc) %>%
    summarise(
      prob_median = median(prob),
      prob_lo     = ggdist::hdi(prob, .width = 0.95)[1],
      prob_hi     = ggdist::hdi(prob, .width = 0.95)[2],
      .groups     = "drop"
    ) %>%
    mutate(
      concentration = to_conc(x_sc),
      species       = species_name,
      line_group    = group_name
    )
}

curve_high_dd <- make_group_curve(
  fit_dd, df_model_dd, high_crosses_dd,
  "D. dianthus", "D. dianthus – High", n_grid
)

curve_low_dd <- make_group_curve(
  fit_dd, df_model_dd, low_crosses_dd,
  "D. dianthus", "D. dianthus – Low", n_grid
)

curves_dd <- bind_rows(curve_high_dd, curve_low_dd)
```

prepare observed points for D. dianthus

```{r}
obs_points_dd <- df0_dd %>%
  mutate(
    concentration = to_conc(x_sc),
    fert_prop     = y / n,
    species       = "D. dianthus"
  ) %>%
  left_join(cross_groups_dd %>% select(cross, line_group), by = "cross")
```

Repeat for D. pertusum
prepare data
```{r}
df0_dp <- db %>%
  filter(species == "D. pertusum") %>%
  transmute(
    cross,
    y    = as.numeric(fertilised),
    n    = as.numeric(particles),
    x_sc = as.numeric(conc_log)
  ) %>%
  filter(!is.na(y), !is.na(n), n > 0, is.finite(x_sc))

if (nrow(df0_dp) == 0) stop("No data found for D. pertusum")

never_fert_crosses_dp <- df0_dp %>%
  group_by(cross) %>%
  summarise(total_fert = sum(y), .groups = "drop") %>%
  filter(total_fert == 0) %>%
  pull(cross)

df_model_dp <- df0_dp %>%
  filter(!cross %in% never_fert_crosses_dp)
```

Fit D. pertusum model

```{r}
priors_dp <- c(
  prior(normal(0, 5), class = "Intercept"),
  prior(normal(0, 3), class = "b"),
  prior(exponential(1), class = "sd", group = "cross"),
  prior(lkj(2), class = "cor")
)

fit_dp <- brm(
  y | trials(n) ~ 1 + x_sc + (1 + x_sc | cross),
  data    = df_model_dp,
  family  = binomial(link = "logit"),
  prior   = priors_dp,
  chains  = chains,
  iter    = iter,
  warmup  = warmup,
  cores   = cores,
  control = list(adapt_delta = 0.97, max_treedepth = 12),
  seed    = seed,
  backend = "cmdstanr"
)
```

Diagnostics

```{r}
summary(fit_dp)
max(rhat(fit_dp), na.rm = TRUE)
min(neff_ratio(fit_dp), na.rm = TRUE)

plot(fit_dp)

pp_check(fit_dp)
pp_check(fit_dp, type = "bars", ndraws = 50)
```

Benchmark classification for D. pertusum

```{r}
c_bench_dp <- 1e6

newdata_bench_dp <- tibble(
  cross = unique(df_model_dp$cross),
  x_sc  = log10(c_bench_dp + 1),
  n     = 1L
)

bench_draws_dp <- add_epred_draws(
  fit_dp,
  newdata = newdata_bench_dp,
  re_formula = NULL
)

prob_col_dp <- if (".epred" %in% names(bench_draws_dp)) ".epred" else ".value"

bench_summary_dp <- bench_draws_dp %>%
  group_by(cross) %>%
  summarise(
    p_reach_target = mean(!!sym(prob_col_dp) >= target_prob),
    prob_median    = median(!!sym(prob_col_dp)),
    .groups = "drop"
  )

cross_groups_dp <- bench_summary_dp %>%
  mutate(
    line_group = if_else(
      p_reach_target >= p_cut,
      "D. pertusum – High",
      "D. pertusum – Low"
    ),
    species = "D. pertusum"
  )

high_crosses_dp <- cross_groups_dp %>%
  filter(line_group == "D. pertusum – High") %>%
  pull(cross)

low_crosses_dp <- cross_groups_dp %>%
  filter(line_group == "D. pertusum – Low") %>%
  pull(cross)
```

Group-average curves for D. pertusum

```{r}
curve_high_dp <- make_group_curve(
  fit_dp, df_model_dp, high_crosses_dp,
  "D. pertusum", "D. pertusum – High", n_grid
)

curve_low_dp <- make_group_curve(
  fit_dp, df_model_dp, low_crosses_dp,
  "D. pertusum", "D. pertusum – Low", n_grid
)

curves_dp <- bind_rows(curve_high_dp, curve_low_dp)
```

Observed data for D. pertusum plot

```{r}
obs_points_dp <- df0_dp %>%
  mutate(
    concentration = to_conc(x_sc),
    fert_prop     = y / n,
    species       = "D. pertusum"
  ) %>%
  left_join(cross_groups_dp %>% select(cross, line_group), by = "cross") %>%
  mutate(
    line_group = if_else(
      cross %in% never_fert_crosses_dp,
      "D. pertusum – Never fertilised",
      line_group
    )
  )
```

Build combined Figure 4

```{r}
curves_all <- bind_rows(curves_dd, curves_dp)
obs_all    <- bind_rows(obs_points_dd, obs_points_dp)

never_lines_dp <- tibble()
if (length(never_fert_crosses_dp) > 0) {
  never_lines_dp <- df0_dp %>%
    filter(cross %in% never_fert_crosses_dp) %>%
    summarise(
      x_start = to_conc(min(x_sc, na.rm = TRUE)),
      x_end   = to_conc(max(x_sc, na.rm = TRUE))
    ) %>%
    mutate(line_group = "D. pertusum – Never fertilised")
}

pal_all <- c(
  "D. dianthus – High"             = "#9467bd",
  "D. dianthus – Low"              = "#674884",
  "D. pertusum – High"             = "#f3c300",
  "D. pertusum – Low"              = "#614e00",
  "D. pertusum – Never fertilised" = "#333333"
)

curves_all <- curves_all %>%
  mutate(concentration = pmax(concentration, .Machine$double.eps))

obs_all <- obs_all %>%
  mutate(concentration = pmax(concentration, .Machine$double.eps))

p_both <- ggplot() +
  geom_ribbon(
    data = curves_all,
    aes(x = concentration, ymin = prob_lo, ymax = prob_hi, fill = line_group),
    alpha = 0.18, colour = NA
  ) +
  geom_line(
    data = curves_all,
    aes(x = concentration, y = prob_median, colour = line_group),
    linewidth = 1
  ) +
  geom_point(
    data = obs_all,
    aes(x = concentration, y = fert_prop, colour = line_group),
    size = 1.8, alpha = 0.75
  ) +
  {if (nrow(never_lines_dp) > 0)
    geom_segment(
      data = never_lines_dp,
      aes(x = x_start, xend = x_end, y = 0, yend = 0, colour = line_group),
      linetype = 2, linewidth = 1
    )
  } +
  scale_x_log10(
    breaks = 10^(0:7),
    labels = trans_format("log10", math_format(10^.x))
  ) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    limits = c(0, 1)
  ) +
  coord_cartesian(xlim = c(1, 1e7)) +
  scale_colour_manual(values = pal_all, name = "Species × Group") +
  scale_fill_manual(values = pal_all, guide = "none") +
  labs(
    x = "Sperm concentration (log10 scale)",
    y = "Fertilisation proportion"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")

p_both
```
Add C50 values to the figure

```{r}
c50_df <- curves_all %>%
  mutate(x_sc = log10(concentration + 1)) %>%
  group_by(line_group) %>%
  arrange(x_sc, .by_group = TRUE) %>%
  mutate(
    p1 = prob_median,
    p2 = lead(prob_median),
    x1 = x_sc,
    x2 = lead(x_sc),
    crosses = (p1 <= 0.5 & p2 >= 0.5) | (p1 >= 0.5 & p2 <= 0.5)
  ) %>%
  filter(crosses) %>%
  slice(1) %>%
  summarise(
    x50_sc = x1 + (0.5 - p1) * (x2 - x1) / (p2 - p1),
    conc50 = 10^x50_sc - 1,
    .groups = "drop"
  ) %>%
  filter(is.finite(conc50), conc50 > 0)

p_both_c50 <- p_both +
  geom_segment(
    data = c50_df,
    aes(x = conc50, xend = conc50, y = 0, yend = 0.5, colour = line_group),
    linetype = "dashed",
    linewidth = 0.6,
    show.legend = FALSE
  ) +
  geom_point(
    data = c50_df,
    aes(x = conc50, y = 0.5, colour = line_group),
    shape = 21,
    fill = "white",
    size = 2.6,
    stroke = 0.8,
    show.legend = FALSE
  )

p_both_c50
```
export figure

```{r}
width_mm  <- 174
height_mm <- 115

quartz(
  type = "pdf",
  file = "../Figures/Figure_dose_response_C50.pdf",
  width = width_mm / 25.4,
  height = height_mm / 25.4
)
print(p_both_c50)
dev.off()


ggsave(
  filename = "../Figures/Figure_dose_response_C50.tiff",
  plot     = p_both_c50,
  device   = "tiff",
  width    = width_mm / 25.4,
  height   = height_mm / 25.4,
  units    = "in",
  dpi      = 600,
  compression = "lzw"
)
```

Export cross-classification summary tables

```{r}
write.csv(cross_groups_dd, "../Tables/cross_groups_dianthus.csv", row.names = FALSE)
write.csv(cross_groups_dp, "../Tables/cross_groups_pertusum.csv", row.names = FALSE)
write.csv(c50_df, "../Tables/c50_summary.csv", row.names = FALSE)
```

Suplementary table S2

```{r}

library(dplyr)
library(stringr)
library(readr)

# ---------- Helper: consistent cross labels ----------
format_cross_id <- function(cross, species) {
  cross_std <- cross %>%
    str_trim() %>%
    str_replace_all("/", "-") %>%
    str_replace_all("\\s+", "")
  
  case_when(
    species == "D. dianthus" ~ cross_std %>%
      str_replace_all("\\b9\\.1\\b", "Dd09") %>%
      str_replace_all("\\b(\\d+)\\b", "Dd\\1"),
    species == "D. pertusum" ~ cross_std,
    TRUE ~ cross_std
  )
}

# ---------- Build cross-class lookup ----------
cross_groups_dd2 <- cross_groups_dd %>%
  mutate(
    fert_class = if_else(str_detect(line_group, "High"), "High", "Low"),
    benchmark_label = "10^4"
  ) %>%
  select(cross, species, fert_class, benchmark_label, p_reach_target, prob_median)

if (length(never_fert_crosses_dd) > 0) {
  cross_groups_dd2 <- bind_rows(
    cross_groups_dd2,
    tibble(
      cross = never_fert_crosses_dd,
      species = "D. dianthus",
      fert_class = "Never",
      benchmark_label = "10^4",
      p_reach_target = NA_real_,
      prob_median = NA_real_
    )
  )
}

cross_groups_dp2 <- cross_groups_dp %>%
  mutate(
    fert_class = if_else(str_detect(line_group, "High"), "High", "Low"),
    benchmark_label = "10^6"
  ) %>%
  select(cross, species, fert_class, benchmark_label, p_reach_target, prob_median)

if (length(never_fert_crosses_dp) > 0) {
  cross_groups_dp2 <- bind_rows(
    cross_groups_dp2,
    tibble(
      cross = never_fert_crosses_dp,
      species = "D. pertusum",
      fert_class = "Never",
      benchmark_label = "10^6",
      p_reach_target = NA_real_,
      prob_median = NA_real_
    )
  )
}

cross_groups_all <- bind_rows(cross_groups_dd2, cross_groups_dp2) %>%
  distinct(cross, species, .keep_all = TRUE)

# ---------- Build supplementary table ----------
table_s2 <- db %>%
  mutate(
    date = as.Date(as.character(date), format = "%d-%m-%Y"),
    cross_label = format_cross_id(cross, species),
    egg_age_display = if_else(egg_age_imputed, paste0(egg_age.hours, "*"), as.character(egg_age.hours)),
    sperm_age_display = if_else(sperm_age_imputed, paste0(sperm_age.hours, "*"), as.character(sperm_age.hours))
  ) %>%
  left_join(
    cross_groups_all,
    by = c("cross", "species")
  ) %>%
  group_by(date, species, cross, cross_label, fert_class, benchmark_label) %>%
  summarise(
    egg_age_h = first(egg_age_display),
    sperm_age_h = first(sperm_age_display),
    p_reach_target = first(p_reach_target),
    prob_median = first(prob_median),
    .groups = "drop"
  ) %>%
  mutate(
    Date = format(date, "%y-%m-%d"),
    Species = species,
    Cross = cross_label,
    `Fert. class` = fert_class,
    `Egg age (h)` = egg_age_h,
    `Sperm age (h)` = sperm_age_h,
    `Benchmark label` = benchmark_label,
    `P(fertilisation probability ≥ 0.5 at benchmark)` =
      if_else(is.na(p_reach_target), NA_character_, format(round(p_reach_target, 3), nsmall = 3)),
    `Predicted fertilisation at benchmark)` =
      if_else(is.na(prob_median), NA_character_, format(round(prob_median, 3), nsmall = 3))
  ) %>%
  select(
    Date,
    Species,
    Cross,
    `Fert. class`,
    `Egg age (h)`,
    `Sperm age (h)`,
    `Benchmark label`,
    `P(fertilisation probability ≥ 0.5 at benchmark)`,
    `Predicted fertilisation at benchmark)`
  ) %>%
  arrange(Species, Date, Cross)

table_s2 <- table_s2 %>%
  rename(`Predicted fertilisation at benchmark` = `Predicted fertilisation at benchmark)`)
table_s2
```

Suplementary figure
Experiment table
```{r}
# ==========================================================
# Create experiment summary table
# Used for supplementary table and supplementary figure
# ==========================================================

cross_groups_dd2 <- cross_groups_dd %>%
  mutate(fert_class = if_else(str_detect(line_group, "High"), "High", "Low")) %>%
  select(cross, species, fert_class, p_reach_target, prob_median)

if (length(never_fert_crosses_dd) > 0) {
  cross_groups_dd2 <- bind_rows(
    cross_groups_dd2,
    tibble(
      cross = never_fert_crosses_dd,
      species = "D. dianthus",
      fert_class = "Never",
      p_reach_target = NA_real_,
      prob_median = NA_real_
    )
  )
}

cross_groups_dp2 <- cross_groups_dp %>%
  mutate(fert_class = if_else(str_detect(line_group, "High"), "High", "Low")) %>%
  select(cross, species, fert_class, p_reach_target, prob_median)

if (length(never_fert_crosses_dp) > 0) {
  cross_groups_dp2 <- bind_rows(
    cross_groups_dp2,
    tibble(
      cross = never_fert_crosses_dp,
      species = "D. pertusum",
      fert_class = "Never",
      p_reach_target = NA_real_,
      prob_median = NA_real_
    )
  )
}

cross_groups_all <- bind_rows(cross_groups_dd2, cross_groups_dp2) %>%
  distinct(cross, species, .keep_all = TRUE)

experiment_table <- db %>%
  select(any_of(c(
    "date", "cross", "species",
    "egg_age.hours", "sperm_age.hours",
    "egg_age_imputed", "sperm_age_imputed",
    "concentration", "conc_log",
    "particles", "fertilised"
  ))) %>%
  mutate(
    date = as.Date(as.character(date), format = "%d-%m-%Y")
  ) %>%
  left_join(
    cross_groups_all %>% select(cross, species, fert_class, p_reach_target, prob_median),
    by = c("cross", "species")
  ) %>%
  group_by(date, species, cross, fert_class) %>%
  summarise(
    egg_age_h   = round(median(egg_age.hours, na.rm = TRUE), 1),
    sperm_age_h = round(median(sperm_age.hours, na.rm = TRUE), 1),
    egg_age_imputed   = any(egg_age_imputed, na.rm = TRUE),
    sperm_age_imputed = any(sperm_age_imputed, na.rm = TRUE),
    concentrations_tested = {
      conc_vals <- concentration[!is.na(concentration)]
      if (length(conc_vals) > 0) {
        paste(sort(unique(conc_vals)), collapse = ", ")
      } else {
        paste(signif(sort(unique(10^conc_log - 1)), 3), collapse = ", ")
      }
    },
    total_particles  = sum(particles, na.rm = TRUE),
    total_fertilised = sum(fertilised, na.rm = TRUE),
    overall_fert_prop = round(total_fertilised / total_particles, 3),
    p_reach_target = round(first(p_reach_target), 3),
    prob_median    = round(first(prob_median), 3),
    .groups = "drop"
  ) %>%
  mutate(
    benchmark_label = case_when(
      species == "D. dianthus" ~ "1e4",
      species == "D. pertusum" ~ "1e6",
      TRUE ~ NA_character_
    )
  ) %>%
  arrange(species, date, fert_class, cross)

experiment_table
```

```{r}
write.csv(table_s2, "../Tables/Table_S2_fertilisation_summary.csv", row.names = FALSE)

write.table(
  table_s2,
  file = "../Tables/Table_S2_fertilisation_summary.txt",
  sep = "\t",
  row.names = FALSE,
  quote = FALSE,
  na = ""
)
```

Supplementary figure

```{r}
## Supplementary Figure S1
## Sperm concentration ranges tested per cross

# ----------------------------------------------------------
# Helper: consistent display labels for crosses
# ----------------------------------------------------------
format_cross_id <- function(cross, species) {
  cross_std <- cross %>%
    stringr::str_trim() %>%
    stringr::str_replace_all("/", "-") %>%
    stringr::str_replace_all("\\s+", "")
  
  dplyr::case_when(
    species == "D. dianthus" ~ cross_std %>%
      stringr::str_replace_all("\\b9\\.1\\b", "Dd09") %>%
      stringr::str_replace_all("\\b(\\d+)\\b", "Dd\\1"),
    species == "D. pertusum" ~ cross_std,
    TRUE ~ cross_std
  )
}

# ----------------------------------------------------------
# Prepare experiment table for plotting
# Requires: experiment_table
# ----------------------------------------------------------
experiment_table_fig <- experiment_table %>%
  mutate(
    date = as.Date(date),
    cross_label = format_cross_id(cross, species),
    exp_id = paste(format(date, "%Y-%m-%d"), cross_label),
    fert_class = factor(fert_class, levels = c("High", "Low", "Never"))
  )

# ----------------------------------------------------------
# Expand concentration list to long format
# ----------------------------------------------------------
df_long <- experiment_table_fig %>%
  mutate(
    conc_list = stringr::str_split(concentrations_tested, "\\s*,\\s*")
  ) %>%
  tidyr::unnest(conc_list) %>%
  mutate(
    concentration = as.numeric(conc_list),
    conc_log10 = log10(concentration + 1)
  ) %>%
  filter(is.finite(conc_log10))

# ----------------------------------------------------------
# Min-max concentration range per experiment
# ----------------------------------------------------------
df_rng <- df_long %>%
  group_by(species, exp_id, date, cross_label, fert_class) %>%
  summarise(
    min_log = min(conc_log10, na.rm = TRUE),
    max_log = max(conc_log10, na.rm = TRUE),
    .groups = "drop"
  )

# ----------------------------------------------------------
# Order by date within species
# ----------------------------------------------------------
df_rng <- df_rng %>%
  group_by(species) %>%
  arrange(desc(date), cross_label, .by_group = TRUE) %>%
  mutate(exp_id_plot = factor(exp_id, levels = unique(exp_id))) %>%
  ungroup()

df_long <- df_long %>%
  left_join(
    df_rng %>% select(species, exp_id, exp_id_plot),
    by = c("species", "exp_id")
  )

# ----------------------------------------------------------
# Build the plot
# ----------------------------------------------------------
p_fig_s1 <- ggplot() +
  geom_segment(
    data = df_rng,
    aes(
      x = min_log, xend = max_log,
      y = exp_id_plot, yend = exp_id_plot,
      colour = fert_class
    ),
    linewidth = 1.2
  ) +
  geom_point(
    data = df_long,
    aes(
      x = conc_log10,
      y = exp_id_plot,
      colour = fert_class
    ),
    shape = 124,
    size = 4,
    alpha = 0.9
  ) +
  facet_grid(
    species ~ .,
    scales = "free_y",
    space = "free_y",
    switch = "y"
  ) +
  scale_colour_manual(
    values = c(
      High  = "#4daf4a",
      Low   = "#377eb8",
      Never = "#e41a1c"
    ),
    drop = FALSE
  ) +
  labs(
    x = expression(log[10]~"(sperm concentration)"),
    y = NULL,
    colour = "Fert class",
    title = "Sperm concentration ranges tested per cross"
  ) +
  theme_bw(base_size = 11) +
  theme(
    legend.position = "bottom",
    panel.grid.major.y = element_blank(),
    strip.background = element_rect(fill = "grey90"),
    strip.placement = "outside"
  )

p_fig_s1
```

export

```{r}
width_mm  <- 174
height_mm <- 115

quartz(
  type = "pdf",
  file = "../Figures/Figure_S1_concentration_ranges.pdf",
  width = width_mm / 25.4,
  height = height_mm / 25.4
)
print(p_fig_s1)
dev.off()

ggsave(
  filename = "../Figures/Figure_S1_concentration_ranges.tiff",
  plot = p_fig_s1,
  device = "tiff",
  width = width_mm / 25.4,
  height = height_mm / 25.4,
  units = "in",
  dpi = 600,
  compression = "lzw"
)
```


