---
title: "Motility_analysis_publication_2026"
author: "Diego Moreno"
date: "2026-02-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

Load packages and read data

```{r}
library(dplyr)
library(ggplot2)
library(brms)
library(patchwork)
library(tidyr)

db <- read.delim("../Data/analysis_input_sperm_motility.txt") %>%
  mutate(
    Species = factor(Species, levels = c("D. pertusum", "D. dianthus"))
  )
head(db)
```
##  Create time bins for summary tables
```{r}
db_binned <- db %>%
  mutate(
    time_bin = cut(
      time_after_spawning_h,
      breaks = c(0, 5, 10, 15, 20, 25, 30, 35, 40),
      labels = c("0-5", "5-10", "10-15", "15-20", "20-25", "25-30", "30-35", "35-40"),
      include.lowest = TRUE,
      right = FALSE
    )
  )
```

Define priors

```{r}
priors <- c(
  prior(normal(0, 1.5), class = "b"),
  prior(exponential(1), class = "sd"),
  prior(gamma(2, 0.1), class = "phi")
)
```

Prepare data for progressive model
```{r}
db_progressive <- db %>%
  filter(
    !is.na(Percent_Progressive),
    !is.na(time_after_spawning_h),
    !is.na(Species),
    !is.na(male)
  ) %>%
  mutate(
    prop_progressive = Percent_Progressive / 100,
    prop_progressive = pmin(pmax(prop_progressive, 0.001), 0.999),
    Species = factor(Species, levels = c("D. pertusum", "D. dianthus")),
    male = factor(male)
  )
```

model progressive

```{r}
mod_bayes_progressive_smooth <- brm(
  formula = prop_progressive ~ Species + s(time_after_spawning_h, by = Species, k = 3) + (1 | male),
  data = db_progressive,
  family = Beta(),
  prior = priors,
  chains = 4,
  cores = 4,
  iter = 4000,
  warmup = 2000,
  seed = 123,
  control = list(adapt_delta = 0.999, max_treedepth = 15)
)
```
summary
```{r}
summary(mod_bayes_progressive_smooth)


```


Prepare data for motile

```{r}
db_motile <- db %>%
  filter(
    !is.na(Percent_Motile),
    !is.na(time_after_spawning_h),
    !is.na(Species),
    !is.na(male)
  ) %>%
  mutate(
    prop_motile = Percent_Motile / 100,
    prop_motile = pmin(pmax(prop_motile, 0.001), 0.999),
    Species = factor(Species, levels = c("D. pertusum", "D. dianthus")),
    male = factor(male)
  )
```

Fit motile smooth model

```{r}
mod_bayes_motile_smooth <- brm(
  formula = prop_motile ~ Species + s(time_after_spawning_h, by = Species, k = 3) + (1 | male),
  data = db_motile,
  family = Beta(),
  prior = priors,
  chains = 4,
  cores = 4,
  iter = 4000,
  warmup = 2000,
  seed = 123,
  control = list(adapt_delta = 0.999, max_treedepth = 15)
)
```
Summary

```{r}
summary(mod_bayes_motile_smooth)
```

Diagnostics

```{r}
summary(mod_bayes_progressive_smooth)
summary(mod_bayes_motile_smooth)

plot(mod_bayes_progressive_smooth)
plot(mod_bayes_motile_smooth)

pp_check(mod_bayes_progressive_smooth)
pp_check(mod_bayes_motile_smooth)
```

Prediction grid for the progressive model

```{r}
newdat_progressive <- expand.grid(
  time_after_spawning_h = seq(
    min(db_progressive$time_after_spawning_h),
    max(db_progressive$time_after_spawning_h),
    length.out = 200
  ),
  Species = levels(db_progressive$Species)
)

pred_progressive <- fitted(
  mod_bayes_progressive_smooth,
  newdata = newdat_progressive,
  re_formula = NA
)

plotdat_progressive <- cbind(newdat_progressive, as.data.frame(pred_progressive)) %>%
  mutate(
    Estimate = Estimate * 100,
    Q2.5 = Q2.5 * 100,
    Q97.5 = Q97.5 * 100
  )
```

Prediction grid for the motile model

```{r}
newdat_motile <- expand.grid(
  time_after_spawning_h = seq(
    min(db_motile$time_after_spawning_h),
    max(db_motile$time_after_spawning_h),
    length.out = 200
  ),
  Species = levels(db_motile$Species)
)

pred_motile <- fitted(
  mod_bayes_motile_smooth,
  newdata = newdat_motile,
  re_formula = NA
)

plotdat_motile <- cbind(newdat_motile, as.data.frame(pred_motile)) %>%
  mutate(
    Estimate = Estimate * 100,
    Q2.5 = Q2.5 * 100,
    Q97.5 = Q97.5 * 100
  )
```

plot progressive model
```{r}
p_progressive_bayes_smooth <- ggplot(
  plotdat_progressive,
  aes(
    x = time_after_spawning_h,
    y = Estimate,
    colour = Species,
    fill = Species,
    group = Species
  )
) +
  geom_point(
    data = db_progressive,
    aes(x = time_after_spawning_h, y = Percent_Progressive, colour = Species),
    alpha = 0.25,
    size = 2,
    inherit.aes = FALSE
  ) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 1.2) +
  scale_colour_manual(
    values = c("D. pertusum" = "#f3c300", "D. dianthus" = "#9467bd"),
    labels = c(
      expression(italic("D. pertusum")),
      expression(italic("D. dianthus"))
    )
  ) +
  scale_fill_manual(
    values = c("D. pertusum" = "#f3c300", "D. dianthus" = "#9467bd"),
    labels = c(
      expression(italic("D. pertusum")),
      expression(italic("D. dianthus"))
    )
  ) +
  labs(
    x = "Time after spawning (hrs)",
    y = "Percent Progressive (%)",
    colour = "Species",
    fill = "Species"
  ) +
  coord_cartesian(ylim = c(0, 30)) +
  theme_minimal(base_size = 12) +
  theme(
    axis.line = element_line(linewidth = 0.7, colour = "black"),
    axis.ticks = element_line(linewidth = 0.7),
    axis.ticks.length = unit(0.1, "cm")
  )
p_progressive_bayes_smooth
```
plot motile model

```{r}
p_motile_bayes_smooth <- ggplot(
  plotdat_motile,
  aes(
    x = time_after_spawning_h,
    y = Estimate,
    colour = Species,
    fill = Species,
    group = Species
  )
) +
  geom_point(
    data = db_motile,
    aes(x = time_after_spawning_h, y = Percent_Motile, colour = Species),
    alpha = 0.25,
    size = 2,
    inherit.aes = FALSE
  ) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 1.2) +
  scale_colour_manual(
    values = c("D. pertusum" = "#f3c300", "D. dianthus" = "#9467bd"),
    labels = c(
      expression(italic("D. pertusum")),
      expression(italic("D. dianthus"))
    )
  ) +
  scale_fill_manual(
    values = c("D. pertusum" = "#f3c300", "D. dianthus" = "#9467bd"),
    labels = c(
      expression(italic("D. pertusum")),
      expression(italic("D. dianthus"))
    )
  ) +
  labs(
    x = "Time after spawning (hrs)",
    y = "Percent Motile (%)",
    colour = "Species",
    fill = "Species"
  ) +
  coord_cartesian(ylim = c(0, 35)) +
  theme_minimal(base_size = 12) +
  theme(
    axis.line = element_line(linewidth = 0.7, colour = "black"),
    axis.ticks = element_line(linewidth = 0.7),
    axis.ticks.length = unit(0.1, "cm")
  )
p_motile_bayes_smooth
```

Combine for figure 2

```{r}
fig_motility_time <- (p_motile_bayes_smooth + p_progressive_bayes_smooth) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "a", tag_prefix = "(", tag_suffix = ")")

fig_motility_time
```
Save figure

```{r}
ggsave(
  filename = "../Figures/Fig_motility_time.tiff",
  plot = fig_motility_time,
  device = "tiff",
  width = 180,
  height = 90,
  units = "mm",
  dpi = 600,
  compression = "lzw",
  bg = "white"
)

ggsave(
  filename = "../Figures/Fig_motility_time.pdf",
  plot = fig_motility_time,
  width = 180,
  height = 90,
  units = "mm",
  bg = "white"
)
```

Build table for summaries

```{r}
all_bin_levels <- c("0-5", "5-10", "10-15", "15-20", "20-25", "25-30", "30-35", "35-40")

db_binned <- db_binned %>%
  mutate(time_bin = factor(time_bin, levels = all_bin_levels))

tab_B_full <- db_binned %>%
  filter(!is.na(time_bin)) %>%
  group_by(Species, time_bin) %>%
  summarise(
    vap_mean = mean(Avg_Motile_VAP, na.rm = TRUE),
    vap_sd   = sd(Avg_Motile_VAP, na.rm = TRUE),
    vap_n    = sum(!is.na(Avg_Motile_VAP)),
    vap_se   = vap_sd / sqrt(vap_n),
    vap_l    = ifelse(vap_n > 1, vap_mean - qt(0.975, df = vap_n - 1) * vap_se, NA_real_),
    vap_u    = ifelse(vap_n > 1, vap_mean + qt(0.975, df = vap_n - 1) * vap_se, NA_real_),

    pvap_mean = mean(Avg_Prog_Motile_VAP, na.rm = TRUE),
    pvap_sd   = sd(Avg_Prog_Motile_VAP, na.rm = TRUE),
    pvap_n    = sum(!is.na(Avg_Prog_Motile_VAP)),
    pvap_se   = pvap_sd / sqrt(pvap_n),
    pvap_l    = ifelse(pvap_n > 1, pvap_mean - qt(0.975, df = pvap_n - 1) * pvap_se, NA_real_),
    pvap_u    = ifelse(pvap_n > 1, pvap_mean + qt(0.975, df = pvap_n - 1) * pvap_se, NA_real_),
    .groups = "drop"
  ) %>%
  mutate(
    `Motile VAP` = ifelse(is.na(vap_l), NA_character_,
                          sprintf("%.2f (%.2f–%.2f)", vap_mean, vap_l, vap_u)),
    `Prog. VAP`  = ifelse(is.na(pvap_l), NA_character_,
                          sprintf("%.2f (%.2f–%.2f)", pvap_mean, pvap_l, pvap_u))
  ) %>%
  select(Species, time_bin, `Motile VAP`, `Prog. VAP`) %>%
  pivot_wider(
    names_from = Species,
    values_from = c(`Motile VAP`, `Prog. VAP`),
    names_glue = "{Species} {.value}"
  ) %>%
  arrange(time_bin) %>%
  mutate(across(-time_bin, ~replace_na(.x, "—"))) %>%
  rename(Time = time_bin)

tab_B_full
```

Table with measurements in both species

```{r}
paper_bins <- c("0-5", "15-20", "25-30")

tab_B_paper <- tab_B_full %>%
  filter(Time %in% paper_bins) %>%
  mutate(Time = factor(Time, levels = paper_bins)) %>%
  arrange(Time)

tab_B_paper
```

export tables

```{r}
write.table(
  tab_B_full,
  "../Tables/TableX_B_velocity_all_bins.txt",
  sep = "\t",
  row.names = FALSE,
  quote = FALSE
)

write.table(
  tab_B_paper,
  "../Tables/TableX_B_velocity_paper.txt",
  sep = "\t",
  row.names = FALSE,
  quote = FALSE
)
```

Plot velocities 

```{r}
## Plot velocity summaries across time bins

velocity_plotdat <- db_binned %>%
  filter(!is.na(time_bin)) %>%
  group_by(Species, time_bin) %>%
  summarise(
    motile_vap_mean = mean(Avg_Motile_VAP, na.rm = TRUE),
    motile_vap_se = sd(Avg_Motile_VAP, na.rm = TRUE) / sqrt(sum(!is.na(Avg_Motile_VAP))),
    prog_vap_mean = mean(Avg_Prog_Motile_VAP, na.rm = TRUE),
    prog_vap_se = sd(Avg_Prog_Motile_VAP, na.rm = TRUE) / sqrt(sum(!is.na(Avg_Prog_Motile_VAP))),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(motile_vap_mean, prog_vap_mean),
    names_to = "Velocity_type",
    values_to = "Mean"
  ) %>%
  mutate(
    SE = ifelse(
      Velocity_type == "motile_vap_mean",
      motile_vap_se,
      prog_vap_se
    ),
    Velocity_type = recode(
      Velocity_type,
      motile_vap_mean = "Motile VAP",
      prog_vap_mean = "Progressive VAP"
    ),
    time_bin = factor(time_bin, levels = all_bin_levels)
  )

p_velocity_time <- ggplot(
  velocity_plotdat,
  aes(
    x = time_bin,
    y = Mean,
    colour = Species,
    group = Species
  )
) +
  geom_errorbar(
    aes(ymin = Mean - SE, ymax = Mean + SE),
    width = 0.15,
    alpha = 0.7
  ) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  facet_wrap(~ Velocity_type, scales = "free_y") +
  scale_colour_manual(
    values = c("D. pertusum" = "#f3c300", "D. dianthus" = "#9467bd"),
    labels = c(
      expression(italic("D. pertusum")),
      expression(italic("D. dianthus"))
    )
  ) +
  labs(
    x = "Time after spawning (hrs)",
    y = "Velocity VAP",
    colour = "Species"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.line = element_line(linewidth = 0.7, colour = "black"),
    axis.ticks = element_line(linewidth = 0.7),
    axis.ticks.length = unit(0.1, "cm")
  )

p_velocity_time
```

Model the velocities 
preparing data
```{r}
## Velocity models and figure

db_vap_motile <- db %>%
  filter(
    !is.na(Avg_Motile_VAP),
    Avg_Motile_VAP > 0,
    !is.na(time_after_spawning_h),
    !is.na(Species),
    !is.na(male)
  ) %>%
  mutate(
    Species = factor(Species, levels = c("D. pertusum", "D. dianthus")),
    male = factor(male)
  )

db_vap_progressive <- db %>%
  filter(
    !is.na(Avg_Prog_Motile_VAP),
    Avg_Prog_Motile_VAP > 0,
    !is.na(time_after_spawning_h),
    !is.na(Species),
    !is.na(male)
  ) %>%
  mutate(
    Species = factor(Species, levels = c("D. pertusum", "D. dianthus")),
    male = factor(male)
  )
```
priors

```{r}
priors_vap <- c(
  prior(normal(0, 1.5), class = "b"),
  prior(exponential(1), class = "sd"),
  prior(exponential(1), class = "sds"),
  prior(exponential(1), class = "shape")
)
```

models

```{r}
mod_bayes_vap_motile <- brm(
  Avg_Motile_VAP ~ Species +
    s(time_after_spawning_h, by = Species, k = 3) +
    (1 | male),
  data = db_vap_motile,
  family = Gamma(link = "log"),
  prior = priors_vap,
  chains = 4,
  cores = 4,
  iter = 4000,
  warmup = 2000,
  seed = 123,
  control = list(adapt_delta = 0.999, max_treedepth = 15)
)

mod_bayes_vap_progressive <- brm(
  Avg_Prog_Motile_VAP ~ Species +
    s(time_after_spawning_h, by = Species, k = 3) +
    (1 | male),
  data = db_vap_progressive,
  family = Gamma(link = "log"),
  prior = priors_vap,
  chains = 4,
  cores = 4,
  iter = 4000,
  warmup = 2000,
  seed = 123,
  control = list(adapt_delta = 0.999, max_treedepth = 15)
)
```

summaries
```{r}
summary(mod_bayes_vap_motile)
summary(mod_bayes_vap_progressive)

plot(mod_bayes_vap_motile)
plot(mod_bayes_vap_progressive)

pp_check(mod_bayes_vap_motile)
pp_check(mod_bayes_vap_progressive)
```

prediction grid for motile VAP

```{r}
## Prediction grid for motile VAP model

newdat_vap_motile <- expand.grid(
  time_after_spawning_h = seq(
    min(db_vap_motile$time_after_spawning_h),
    max(db_vap_motile$time_after_spawning_h),
    length.out = 200
  ),
  Species = levels(db_vap_motile$Species)
)

pred_vap_motile <- fitted(
  mod_bayes_vap_motile,
  newdata = newdat_vap_motile,
  re_formula = NA
)

plotdat_vap_motile <- cbind(
  newdat_vap_motile,
  as.data.frame(pred_vap_motile)
)
```

prediction grid for progressive VAP

```{r}
## Prediction grid for progressive VAP model

# newdat_vap_progressive <- expand.grid(
#   time_after_spawning_h = seq(
#     min(db_vap_progressive$time_after_spawning_h),
#     max(db_vap_progressive$time_after_spawning_h),
#     length.out = 200
#   ),
#   Species = levels(db_vap_progressive$Species)
# )

# prediction grid only for the observations we had 

newdat_vap_progressive <- db_vap_progressive %>%
  group_by(Species) %>%
  summarise(
    min_time = min(time_after_spawning_h, na.rm = TRUE),
    max_time = max(time_after_spawning_h, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  rowwise() %>%
  do(data.frame(
    Species = .$Species,
    time_after_spawning_h = seq(.$min_time, .$max_time, length.out = 200)
  )) %>%
  ungroup() %>%
  mutate(
    Species = factor(Species, levels = levels(db_vap_progressive$Species))
  )

pred_vap_progressive <- fitted(
  mod_bayes_vap_progressive,
  newdata = newdat_vap_progressive,
  re_formula = NA
)

plotdat_vap_progressive <- cbind(
  newdat_vap_progressive,
  as.data.frame(pred_vap_progressive)
)
```

colours and labels

```{r}
## Shared colours and labels

species_cols <- c(
  "D. pertusum" = "#f3c300",
  "D. dianthus" = "#9467bd"
)

species_labs <- c(
  expression(italic("D. pertusum")),
  expression(italic("D. dianthus"))
)
```

plot motile vap

```{r}
## Plot motile VAP model

p_vap_motile <- ggplot(
  plotdat_vap_motile,
  aes(
    x = time_after_spawning_h,
    y = Estimate,
    colour = Species,
    fill = Species,
    group = Species
  )
) +
  geom_point(
    data = db_vap_motile,
    aes(
      x = time_after_spawning_h,
      y = Avg_Motile_VAP,
      colour = Species
    ),
    alpha = 0.25,
    size = 2,
    inherit.aes = FALSE
  ) +
  geom_ribbon(
    aes(ymin = Q2.5, ymax = Q97.5),
    alpha = 0.2,
    colour = NA
  ) +
  geom_line(linewidth = 1.2) +
  scale_colour_manual(
    values = species_cols,
    labels = species_labs
  ) +
  scale_fill_manual(
    values = species_cols,
    labels = species_labs
  ) +
  labs(
    x = "Time after spawning (hrs)",
    y = expression("Motile VAP ("*mu*"m s"^{-1}*")"),
    colour = "Species",
    fill = "Species"
  ) +
  coord_cartesian(
    ylim = c(20,75)
  ) +
  # coord_cartesian(
  #   ylim = c(0, max(plotdat_vap_motile$Q97.5, na.rm = TRUE) * 1.05)
  # ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.line = element_line(linewidth = 0.7, colour = "black"),
    axis.ticks = element_line(linewidth = 0.7),
    axis.ticks.length = unit(0.1, "cm")
  )

p_vap_motile
```

plot progressive VAP

```{r}
## Plot progressive VAP model

p_vap_progressive <- ggplot(
  plotdat_vap_progressive,
  aes(
    x = time_after_spawning_h,
    y = Estimate,
    colour = Species,
    fill = Species,
    group = Species
  )
) +
  geom_point(
    data = db_vap_progressive,
    aes(
      x = time_after_spawning_h,
      y = Avg_Prog_Motile_VAP,
      colour = Species
    ),
    alpha = 0.25,
    size = 2,
    inherit.aes = FALSE
  ) +
  geom_ribbon(
    aes(ymin = Q2.5, ymax = Q97.5),
    alpha = 0.2,
    colour = NA
  ) +
  geom_line(linewidth = 1.2) +
  scale_colour_manual(
    values = species_cols,
    labels = species_labs
  ) +
  scale_fill_manual(
    values = species_cols,
    labels = species_labs
  ) +
  labs(
    x = "Time after spawning (hrs)",
    y = expression("Progressive VAP ("*mu*"m s"^{-1}*")"),
    colour = "Species",
    fill = "Species"
  ) +
  coord_cartesian(
    ylim = c(75,160)
  ) +
  #coord_cartesian(
    #ylim = c(0, max(plotdat_vap_progressive$Q97.5, na.rm = TRUE) * 1.05)
  #) +
  theme_minimal(base_size = 12) +
  theme(
    axis.line = element_line(linewidth = 0.7, colour = "black"),
    axis.ticks = element_line(linewidth = 0.7),
    axis.ticks.length = unit(0.1, "cm")
  )

p_vap_progressive
```

figure 2 expanded

```{r}
fig_motility_time_2 <- (p_motile_bayes_smooth + p_progressive_bayes_smooth)/(p_vap_motile+p_vap_progressive) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "a", tag_prefix = "(", tag_suffix = ")") &
  theme(legend.position = "bottom")

fig_motility_time_2
```

save

```{r}
ggsave(
  filename = "../Figures/Fig2_motility_velocity.tiff",
  plot = fig_motility_time_2,
  device = "tiff",
  width = 174,
  height = 140,
  units = "mm",
  dpi = 600,
  compression = "lzw",
  bg = "white"
)

ggsave(
  filename = "../Figures/Fig2_motility_velocity.pdf",
  plot = fig_motility_time_2,
  width = 174,
  height = 140,
  units = "mm",
  bg = "white"
)
```

