---
title: "Motility_analysis_publication_2026"
author: "Diego Moreno"
date: "2026-02-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

read data

```{r}
db <- read.delim("../Data/analysis_input_sperm_motility.txt") %>%
  mutate(
    Species = factor(Species, levels = c("D. pertusum", "D. dianthus"))
  )
head(db)
```
Binning for trend visualization
```{r}
library(dplyr)
library(ggplot2)

#create bins
db_binned <- db %>%
  mutate(time_bin = cut(time_after_spawning_h, 
                        breaks = c(0, 5, 10, 15, 20, 25, 30, 35, 40), # Define your desired bins
                        labels = c("0-5", "5-10", "10-15", "15-20", "20-25", "25-30", "30-35", "35-40"), # Labels for bins
                        include.lowest = TRUE, right = FALSE)) %>% # include_lowest ensures 0 is included, right=FALSE means [a,b)
  # Convert to factor and reverse levels so earliest times are at the bottom
  mutate(time_bin = factor(time_bin, levels = rev(levels(time_bin))))
```

test plot

```{r}

db_bind_summary_motile <- db_binned %>%
  group_by(Species, time_bin) %>%
  summarise(
    mean_motile = mean(Percent_Motile, na.rm = TRUE),
    sd_motile = sd(Percent_Motile, na.rm = TRUE),
    n = sum(!is.na(Percent_Motile)),
    se = sd_motile / sqrt(n),
    ci_lower = mean_motile - qt(0.975, df = n-1) * se,
    ci_upper = mean_motile + qt(0.975, df = n-1) * se
  )

db_bind_summary_motile <- db_bind_summary_motile %>%
  mutate(time_bin = factor(time_bin, levels = c("0-5", "5-10", "10-15", "15-20", "20-25", "25-30", "30-35", "35-40")))


ggplot(db_bind_summary_motile, aes(x = time_bin, y = mean_motile, group = Species, colour = Species)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
  scale_colour_manual(values = c("D. pertusum" = "#f3c300", "D. dianthus" = "#9467bd")) +
  theme_minimal()
```


function for ploting

```{r}

#labels staggered

time_bin_levels <- c(
  "0-5", "5-10", "10-15", "15-20",
  "20-25", "25-30", "30-35", "35-40"
)

time_bin_labels <- c(
  "0–5\n", 
  "\n5–10", 
  "10–15\n", 
  "\n15–20",
  "20–25\n", 
  "\n25–30", 
  "30–35\n", 
  "\n35–40"
)


plot_time_bin <- function(data, var, y_label) {

  summary_data <- data %>%
    group_by(Species, time_bin) %>%
    summarise(
      n = sum(!is.na(.data[[var]])),
      mean_val = mean(.data[[var]], na.rm = TRUE),
      sd_val   = sd(.data[[var]], na.rm = TRUE),
      se       = ifelse(n > 1, sd_val / sqrt(n), NA_real_),
      ci_lower = ifelse(n > 1, mean_val - qt(0.975, df = n-1) * se, NA_real_),
      ci_upper = ifelse(n > 1, mean_val + qt(0.975, df = n-1) * se, NA_real_),
      .groups = "drop"
    ) %>%
    mutate(time_bin = factor(time_bin, levels = time_bin_levels))

  ggplot(summary_data, aes(x = time_bin, y = mean_val, group = Species, colour = Species)) +
    geom_line(size = 1.2) +
    geom_point(size = 3) +
    geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
    scale_colour_manual(
      values = c("D. pertusum" = "#f3c300", "D. dianthus" = "#9467bd"),
      labels = c(expression(italic("D. pertusum")),
                 expression(italic("D. dianthus")))
    ) +
    scale_x_discrete(labels = time_bin_labels) +
    labs(x = "Time after spawning (hrs)", y = y_label, colour = "Species") +
    theme_minimal(base_size = 12) +
    theme(
      axis.line = element_line(linewidth = 0.7, colour = "black"),
      axis.ticks = element_line(linewidth = 0.7),
      axis.ticks.length = unit(0.1, "cm")
    )
}

```

using the function

```{r}
# Percentages
p_motile <- plot_time_bin(db_binned, "Percent_Motile", "Mean Percent Motile (%)")
p_progressive <- plot_time_bin(db_binned, "Percent_Progressive", "Mean Percent Progressive (%)")
p_total_motile <- plot_time_bin(db_binned, "Percent_Total_Motile", "Mean Percent Total Motile (%)")
p_percent_slow <- plot_time_bin(db_binned, "percent_slow", "Mean Percent Slow (%)")

# Average Motile velocities
p_avg_motile_vcl <- plot_time_bin(db_binned, "Avg_Motile_VCL", "Mean Motile VCL (µm/s)")
p_avg_motile_vap <- plot_time_bin(db_binned, "Avg_Motile_VAP", "Mean Motile VAP (µm/s)")
p_avg_motile_vsl <- plot_time_bin(db_binned, "Avg_Motile_VSL", "Mean Motile VSL (µm/s)")

# Average Progressive Motile velocities
p_avg_prog_motile_vcl <- plot_time_bin(db_binned, "Avg_Prog_Motile_VCL", "Mean Progressive Motile VCL (µm/s)")
p_avg_prog_motile_vap <- plot_time_bin(db_binned, "Avg_Prog_Motile_VAP", "Mean Progressive Motile VAP (µm/s)")
p_avg_prog_motile_vsl <- plot_time_bin(db_binned, "Avg_Prog_Motile_VSL", "Mean Progressive Motile VSL (µm/s)")

# Average Total Motile velocities
p_avg_tot_motile_vcl <- plot_time_bin(db_binned, "Avg_Tot_Motile_VCL", "Mean Total Motile VCL (µm/s)")
p_avg_tot_motile_vap <- plot_time_bin(db_binned, "Avg_Tot_Motile_VAP", "Mean Total Motile VAP (µm/s)")
p_avg_tot_motile_vsl <- plot_time_bin(db_binned, "Avg_Tot_Motile_VSL", "Mean Total Motile VSL (µm/s)")

#print
print(p_motile)
print(p_progressive)
print(p_total_motile)
print(p_percent_slow)

print(p_avg_motile_vap)
print(p_avg_prog_motile_vap)
print(p_avg_tot_motile_vap)
```

```{r}
db_binned %>% filter(Species == "D. dianthus") %>% filter(time_after_spawning_h > 25)
```

make the plot for export

```{r}
library(patchwork)

p_motile + p_progressive + plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "a", tag_prefix = "(", tag_suffix = ")")
```

testing a smooth line

```{r}
library(dplyr)
library(ggplot2)
library(brms)
db2 <- db %>%
  mutate(
    prop_progressive = pmin(pmax(Percent_Progressive / 100, 0.001), 0.999)
  )


mod_bayes <- brm(
  prop_progressive ~ Species + s(time_after_spawning_h, by = Species, k = 4),
  data = db2,
  family = Beta(),
  chains = 4,
  cores = 4,
  iter = 4000,
  warmup = 2000,
  control = list(adapt_delta = 0.99, max_treedepth = 12)
)

newdat <- expand.grid(
  time_after_spawning_h = seq(0, 40, length.out = 100),
  Species = unique(db2$Species)
)

pred <- fitted(mod_bayes, newdata = newdat)

plotdat <- cbind(newdat, as.data.frame(pred)) %>%
  mutate(
    Estimate = Estimate * 100,
    Q2.5 = Q2.5 * 100,
    Q97.5 = Q97.5 * 100
  )

ggplot(plotdat, aes(x = time_after_spawning_h, y = Estimate, colour = Species)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5, fill = Species), alpha = 0.2, colour = NA) +
  labs(
    x = "Time after spawning (hrs)",
    y = "Predicted Percent Progressive (%)"
  ) +
  theme_minimal()
```
```{r}
# Prepare raw data for Bayesian model
db2 <- db %>%
  filter(
    !is.na(Percent_Progressive),
    !is.na(time_after_spawning_h),
    !is.na(Species)
  ) %>%
  mutate(
    prop_progressive = Percent_Progressive / 100,
    prop_progressive = pmin(pmax(prop_progressive, 0.001), 0.999),
    Species = factor(Species)
  )

# Weakly regularising priors
priors <- c(
  prior(normal(0, 1.5), class = "b"),
  prior(exponential(1), class = "sds"),
  prior(gamma(2, 0.1), class = "phi")
)

# Bayesian model on full continuous time data
mod_bayes <- brm(
  formula = prop_progressive ~ Species + s(time_after_spawning_h, by = Species, k = 4),
  data = db2,
  family = Beta(),
  prior = priors,
  chains = 4,
  cores = 4,
  iter = 4000,
  warmup = 2000,
  seed = 123,
  control = list(adapt_delta = 0.99, max_treedepth = 12)
)

# Optional checks
print(summary(mod_bayes))
pp_check(mod_bayes)

# Prediction grid across observed time range
newdat <- expand.grid(
  time_after_spawning_h = seq(
    min(db2$time_after_spawning_h),
    max(db2$time_after_spawning_h),
    length.out = 200
  ),
  Species = levels(db2$Species)
)

pred <- fitted(mod_bayes, newdata = newdat)

plotdat <- cbind(newdat, as.data.frame(pred)) %>%
  mutate(
    Estimate = Estimate * 100,
    Q2.5 = Q2.5 * 100,
    Q97.5 = Q97.5 * 100
  )

# Bayesian smooth plot
p_progressive_bayes <- ggplot(
  plotdat,
  aes(
    x = time_after_spawning_h,
    y = Estimate,
    colour = Species,
    fill = Species,
    group = Species
  )
) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 1.2) +
  scale_colour_manual(
    values = c("D. pertusum" = "#f3c300", "D. dianthus" = "#9467bd"),
    labels = c(expression(italic("D. pertusum")),
               expression(italic("D. dianthus")))
  ) +
  geom_point(
  data = db2,
  aes(x = time_after_spawning_h, y = Percent_Progressive, colour = Species),
  alpha = 0.2,
  inherit.aes = FALSE)+
  scale_fill_manual(
    values = c("D. pertusum" = "#f3c300", "D. dianthus" = "#9467bd"),
    labels = c(expression(italic("D. pertusum")),
               expression(italic("D. dianthus")))
  ) +
  labs(
    x = "Time after spawning (hrs)",
    y = "Predicted Percent Progressive (%)",
    colour = "Species",
    fill = "Species"
  ) +
  coord_cartesian(ylim = c(0, 30)) +
  theme_minimal(base_size = 12) +
  theme(
    axis.line = element_line(linewidth = 0.7, colour = "black"),
    axis.ticks = element_line(linewidth = 0.7),
    axis.ticks.length = unit(0.1, "cm")
  )

print(p_progressive_bayes)

unique(db$Species)
table(db$Species)
```
trying with k=3

```{r}
# Prepare raw data for Bayesian model
db2 <- db %>%
  filter(
    !is.na(Percent_Progressive),
    !is.na(time_after_spawning_h),
    !is.na(Species)
  ) %>%
  mutate(
    prop_progressive = Percent_Progressive / 100,
    prop_progressive = pmin(pmax(prop_progressive, 0.001), 0.999),
    Species = factor(Species)
  )

# Weakly regularising priors
priors <- c(
  prior(normal(0, 1.5), class = "b"),
  prior(exponential(1), class = "sds"),
  prior(gamma(2, 0.1), class = "phi")
)

# Bayesian model on full continuous time data
mod_bayes <- brm(
  formula = prop_progressive ~ Species + s(time_after_spawning_h, by = Species, k = 3)+ (1 | male),
  data = db2,
  family = Beta(),
  prior = priors,
  chains = 4,
  cores = 4,
  iter = 4000,
  warmup = 2000,
  seed = 123,
  control = list(adapt_delta = 0.99, max_treedepth = 12)
)

# Optional checks
print(summary(mod_bayes))
pp_check(mod_bayes)

# Prediction grid across observed time range
newdat <- expand.grid(
  time_after_spawning_h = seq(
    min(db2$time_after_spawning_h),
    max(db2$time_after_spawning_h),
    length.out = 200
  ),
  Species = levels(db2$Species)
)

pred <- fitted(mod_bayes, newdata = newdat, re_formula = NA)

plotdat <- cbind(newdat, as.data.frame(pred)) %>%
  mutate(
    Estimate = Estimate * 100,
    Q2.5 = Q2.5 * 100,
    Q97.5 = Q97.5 * 100
  )

# Bayesian smooth plot
p_progressive_bayes <- ggplot(
  plotdat,
  aes(
    x = time_after_spawning_h,
    y = Estimate,
    colour = Species,
    fill = Species,
    group = Species
  )
) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 1.2) +
  scale_colour_manual(
    values = c("D. pertusum" = "#f3c300", "D. dianthus" = "#9467bd"),
    labels = c(expression(italic("D. pertusum")),
               expression(italic("D. dianthus")))
  ) +
  geom_point(
  data = db2,
  aes(x = time_after_spawning_h, y = Percent_Progressive, colour = Species),
  alpha = 0.2,
  inherit.aes = FALSE)+
  scale_fill_manual(
    values = c("D. pertusum" = "#f3c300", "D. dianthus" = "#9467bd"),
    labels = c(expression(italic("D. pertusum")),
               expression(italic("D. dianthus")))
  ) +
  labs(
    x = "Time after spawning (hrs)",
    y = "Predicted Percent Progressive (%)",
    colour = "Species",
    fill = "Species"
  ) +
  coord_cartesian(ylim = c(0, 30)) +
  theme_minimal(base_size = 12) +
  theme(
    axis.line = element_line(linewidth = 0.7, colour = "black"),
    axis.ticks = element_line(linewidth = 0.7),
    axis.ticks.length = unit(0.1, "cm")
  )

print(p_progressive_bayes)

unique(db$Species)
table(db$Species)
```

another model

```{r}
# Polynomial Bayesian model with male random intercept

library(dplyr)
library(ggplot2)
library(brms)

# Prepare raw data
db2 <- db %>%
  filter(
    !is.na(Percent_Progressive),
    !is.na(time_after_spawning_h),
    !is.na(Species),
    !is.na(male)
  ) %>%
  mutate(
    prop_progressive = Percent_Progressive / 100,
    prop_progressive = pmin(pmax(prop_progressive, 0.001), 0.999),
    Species = factor(Species, levels = c("D. pertusum", "D. dianthus")),
    male = factor(male)
  )

# Priors
priors <- c(
  prior(normal(0, 1.5), class = "b"),
  prior(exponential(1), class = "sd"),
  prior(gamma(2, 0.1), class = "phi")
)

# Fit Bayesian polynomial model
mod_bayes_poly <- brm(
  formula = prop_progressive ~ Species * poly(time_after_spawning_h, 2) + (1 | male),
  data = db2,
  family = Beta(),
  prior = priors,
  chains = 4,
  cores = 4,
  iter = 4000,
  warmup = 2000,
  seed = 123,
  control = list(adapt_delta = 0.99, max_treedepth = 12)
)

# Model summary and posterior predictive check
print(summary(mod_bayes_poly))
pp_check(mod_bayes_poly)

# Prediction grid for population-level predictions
newdat_poly <- expand.grid(
  time_after_spawning_h = seq(
    min(db2$time_after_spawning_h),
    max(db2$time_after_spawning_h),
    length.out = 200
  ),
  Species = levels(db2$Species)
)

pred_poly <- fitted(mod_bayes_poly, newdata = newdat_poly, re_formula = NA)

plotdat_poly <- cbind(newdat_poly, as.data.frame(pred_poly)) %>%
  mutate(
    Estimate = Estimate * 100,
    Q2.5 = Q2.5 * 100,
    Q97.5 = Q97.5 * 100
  )

# Plot with raw points
p_progressive_bayes_poly <- ggplot(
  plotdat_poly,
  aes(
    x = time_after_spawning_h,
    y = Estimate,
    colour = Species,
    fill = Species,
    group = Species
  )
) +
  geom_point(
    data = db2,
    aes(x = time_after_spawning_h, y = Percent_Progressive, colour = Species),
    alpha = 0.25,
    size = 2,
    inherit.aes = FALSE
  ) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 1.2) +
  scale_colour_manual(
    values = c("D. pertusum" = "#f3c300", "D. dianthus" = "#9467bd"),
    labels = c(expression(italic("D. pertusum")),
               expression(italic("D. dianthus")))
  ) +
  scale_fill_manual(
    values = c("D. pertusum" = "#f3c300", "D. dianthus" = "#9467bd"),
    labels = c(expression(italic("D. pertusum")),
               expression(italic("D. dianthus")))
  ) +
  labs(
    x = "Time after spawning (hrs)",
    y = "Predicted Percent Progressive (%)",
    colour = "Species",
    fill = "Species"
  ) +
  coord_cartesian(ylim = c(0, 30)) +
  theme_minimal(base_size = 12) +
  theme(
    axis.line = element_line(linewidth = 0.7, colour = "black"),
    axis.ticks = element_line(linewidth = 0.7),
    axis.ticks.length = unit(0.1, "cm")
  )

print(p_progressive_bayes_poly)
```
for motile sperm as well

```{r}
library(dplyr)
library(ggplot2)
library(brms)
library(patchwork)

priors <- c(
  prior(normal(0, 1.5), class = "b"),
  prior(exponential(1), class = "sd"),
  prior(gamma(2, 0.1), class = "phi")
)

# -----------------------------
# Progressive
# -----------------------------
db_progressive <- db %>%
  filter(
    !is.na(Percent_Progressive),
    !is.na(time_after_spawning_h),
    !is.na(Species),
    !is.na(male)
  ) %>%
  mutate(
    prop_progressive = Percent_Progressive / 100,
    prop_progressive = pmin(pmax(prop_progressive, 0.001), 0.999),
    Species = factor(Species, levels = c("D. pertusum", "D. dianthus")),
    male = factor(male)
  )

mod_bayes_poly_progressive <- brm(
  formula = prop_progressive ~ Species * poly(time_after_spawning_h, 2) + (1 | male),
  data = db_progressive,
  family = Beta(),
  prior = priors,
  chains = 4,
  cores = 4,
  iter = 4000,
  warmup = 2000,
  seed = 123,
  control = list(adapt_delta = 0.99, max_treedepth = 12)
)

newdat_progressive <- expand.grid(
  time_after_spawning_h = seq(
    min(db_progressive$time_after_spawning_h),
    max(db_progressive$time_after_spawning_h),
    length.out = 200
  ),
  Species = levels(db_progressive$Species)
)

pred_progressive <- fitted(
  mod_bayes_poly_progressive,
  newdata = newdat_progressive,
  re_formula = NA
)

plotdat_progressive <- cbind(newdat_progressive, as.data.frame(pred_progressive)) %>%
  mutate(
    Estimate = Estimate * 100,
    Q2.5 = Q2.5 * 100,
    Q97.5 = Q97.5 * 100
  )

p_progressive_bayes_poly <- ggplot(
  plotdat_progressive,
  aes(
    x = time_after_spawning_h,
    y = Estimate,
    colour = Species,
    fill = Species,
    group = Species
  )
) +
  geom_point(
    data = db_progressive,
    aes(x = time_after_spawning_h, y = Percent_Progressive, colour = Species),
    alpha = 0.25,
    size = 2,
    inherit.aes = FALSE
  ) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 1.2) +
  scale_colour_manual(
    values = c("D. pertusum" = "#f3c300", "D. dianthus" = "#9467bd"),
    labels = c(expression(italic("D. pertusum")),
               expression(italic("D. dianthus")))
  ) +
  scale_fill_manual(
    values = c("D. pertusum" = "#f3c300", "D. dianthus" = "#9467bd"),
    labels = c(expression(italic("D. pertusum")),
               expression(italic("D. dianthus")))
  ) +
  labs(
    x = "Time after spawning (hrs)",
    y = "Predicted Percent Progressive (%)",
    colour = "Species",
    fill = "Species"
  ) +
  coord_cartesian(ylim = c(0, 30)) +
  theme_minimal(base_size = 12) +
  theme(
    axis.line = element_line(linewidth = 0.7, colour = "black"),
    axis.ticks = element_line(linewidth = 0.7),
    axis.ticks.length = unit(0.1, "cm")
  )

# -----------------------------
# Motile
# -----------------------------
# -----------------------------
# Motile
# -----------------------------
db_motile <- db %>%
  filter(
    !is.na(Percent_Motile),
    !is.na(time_after_spawning_h),
    !is.na(Species),
    !is.na(male)
  ) %>%
  mutate(
    prop_motile = Percent_Motile / 100,
    prop_motile = pmin(pmax(prop_motile, 0.001), 0.999),
    Species = factor(Species, levels = c("D. pertusum", "D. dianthus")),
    male = factor(male)
  )

mod_bayes_motile_smooth <- brm(
  formula = prop_motile ~ Species + s(time_after_spawning_h, by = Species, k = 3) + (1 | male),
  data = db_motile,
  family = Beta(),
  prior = priors,
  chains = 4,
  cores = 4,
  iter = 4000,
  warmup = 2000,
  seed = 123,
  control = list(adapt_delta = 0.99, max_treedepth = 12)
)

newdat_motile <- expand.grid(
  time_after_spawning_h = seq(
    min(db_motile$time_after_spawning_h),
    max(db_motile$time_after_spawning_h),
    length.out = 200
  ),
  Species = levels(db_motile$Species)
)

pred_motile <- fitted(
  mod_bayes_motile_smooth,
  newdata = newdat_motile,
  re_formula = NA
)

plotdat_motile <- cbind(newdat_motile, as.data.frame(pred_motile)) %>%
  mutate(
    Estimate = Estimate * 100,
    Q2.5 = Q2.5 * 100,
    Q97.5 = Q97.5 * 100
  )

p_motile_bayes_smooth <- ggplot(
  plotdat_motile,
  aes(
    x = time_after_spawning_h,
    y = Estimate,
    colour = Species,
    fill = Species,
    group = Species
  )
) +
  geom_point(
    data = db_motile,
    aes(x = time_after_spawning_h, y = Percent_Motile, colour = Species),
    alpha = 0.25,
    size = 2,
    inherit.aes = FALSE
  ) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 1.2) +
  scale_colour_manual(
    values = c("D. pertusum" = "#f3c300", "D. dianthus" = "#9467bd"),
    labels = c(expression(italic("D. pertusum")),
               expression(italic("D. dianthus")))
  ) +
  scale_fill_manual(
    values = c("D. pertusum" = "#f3c300", "D. dianthus" = "#9467bd"),
    labels = c(expression(italic("D. pertusum")),
               expression(italic("D. dianthus")))
  ) +
  labs(
    x = "Time after spawning (hrs)",
    y = "Predicted Percent Motile (%)",
    colour = "Species",
    fill = "Species"
  ) +
  coord_cartesian(ylim = c(0, 35)) +
  theme_minimal(base_size = 12) +
  theme(
    axis.line = element_line(linewidth = 0.7, colour = "black"),
    axis.ticks = element_line(linewidth = 0.7),
    axis.ticks.length = unit(0.1, "cm")
  )
# -----------------------------
# Combined figure
# -----------------------------
(p_motile_bayes_smooth + p_progressive_bayes_poly) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "a", tag_prefix = "(", tag_suffix = ")")
```

```{r}
library(dplyr)
library(ggplot2)
library(brms)
library(patchwork)
library(grid)

priors <- c(
  prior(normal(0, 1.5), class = "b"),
  prior(exponential(1), class = "sd"),
  prior(gamma(2, 0.1), class = "phi")
)

# -----------------------------
# Progressive
# -----------------------------
db_progressive <- db %>%
  filter(
    !is.na(Percent_Progressive),
    !is.na(time_after_spawning_h),
    !is.na(Species),
    !is.na(male)
  ) %>%
  mutate(
    prop_progressive = Percent_Progressive / 100,
    prop_progressive = pmin(pmax(prop_progressive, 0.001), 0.999),
    Species = factor(Species, levels = c("D. pertusum", "D. dianthus")),
    male = factor(male)
  )

mod_bayes_progressive_smooth <- brm(
  formula = prop_progressive ~ Species + s(time_after_spawning_h, by = Species, k = 3) + (1 | male),
  data = db_progressive,
  family = Beta(),
  prior = priors,
  chains = 4,
  cores = 4,
  iter = 4000,
  warmup = 2000,
  seed = 123,
  control = list(adapt_delta = 0.999, max_treedepth = 15)
)

newdat_progressive <- expand.grid(
  time_after_spawning_h = seq(
    min(db_progressive$time_after_spawning_h),
    max(db_progressive$time_after_spawning_h),
    length.out = 200
  ),
  Species = levels(db_progressive$Species)
)

pred_progressive <- fitted(
  mod_bayes_progressive_smooth,
  newdata = newdat_progressive,
  re_formula = NA
)

plotdat_progressive <- cbind(newdat_progressive, as.data.frame(pred_progressive)) %>%
  mutate(
    Estimate = Estimate * 100,
    Q2.5 = Q2.5 * 100,
    Q97.5 = Q97.5 * 100
  )

p_progressive_bayes_smooth <- ggplot(
  plotdat_progressive,
  aes(
    x = time_after_spawning_h,
    y = Estimate,
    colour = Species,
    fill = Species,
    group = Species
  )
) +
  geom_point(
    data = db_progressive,
    aes(x = time_after_spawning_h, y = Percent_Progressive, colour = Species),
    alpha = 0.25,
    size = 2,
    inherit.aes = FALSE
  ) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 1.2) +
  scale_colour_manual(
    values = c("D. pertusum" = "#f3c300", "D. dianthus" = "#9467bd"),
    labels = c(
      expression(italic("D. pertusum")),
      expression(italic("D. dianthus"))
    )
  ) +
  scale_fill_manual(
    values = c("D. pertusum" = "#f3c300", "D. dianthus" = "#9467bd"),
    labels = c(
      expression(italic("D. pertusum")),
      expression(italic("D. dianthus"))
    )
  ) +
  labs(
    x = "Time after spawning (hrs)",
    y = "Predicted Percent Progressive (%)",
    colour = "Species",
    fill = "Species"
  ) +
  coord_cartesian(ylim = c(0, 30)) +
  theme_minimal(base_size = 12) +
  theme(
    axis.line = element_line(linewidth = 0.7, colour = "black"),
    axis.ticks = element_line(linewidth = 0.7),
    axis.ticks.length = unit(0.1, "cm")
  )

# -----------------------------
# Motile
# -----------------------------
db_motile <- db %>%
  filter(
    !is.na(Percent_Motile),
    !is.na(time_after_spawning_h),
    !is.na(Species),
    !is.na(male)
  ) %>%
  mutate(
    prop_motile = Percent_Motile / 100,
    prop_motile = pmin(pmax(prop_motile, 0.001), 0.999),
    Species = factor(Species, levels = c("D. pertusum", "D. dianthus")),
    male = factor(male)
  )

mod_bayes_motile_smooth <- brm(
  formula = prop_motile ~ Species + s(time_after_spawning_h, by = Species, k = 3) + (1 | male),
  data = db_motile,
  family = Beta(),
  prior = priors,
  chains = 4,
  cores = 4,
  iter = 4000,
  warmup = 2000,
  seed = 123,
  control = list(adapt_delta = 0.999, max_treedepth = 15)
)

newdat_motile <- expand.grid(
  time_after_spawning_h = seq(
    min(db_motile$time_after_spawning_h),
    max(db_motile$time_after_spawning_h),
    length.out = 200
  ),
  Species = levels(db_motile$Species)
)

pred_motile <- fitted(
  mod_bayes_motile_smooth,
  newdata = newdat_motile,
  re_formula = NA
)

plotdat_motile <- cbind(newdat_motile, as.data.frame(pred_motile)) %>%
  mutate(
    Estimate = Estimate * 100,
    Q2.5 = Q2.5 * 100,
    Q97.5 = Q97.5 * 100
  )

p_motile_bayes_smooth <- ggplot(
  plotdat_motile,
  aes(
    x = time_after_spawning_h,
    y = Estimate,
    colour = Species,
    fill = Species,
    group = Species
  )
) +
  geom_point(
    data = db_motile,
    aes(x = time_after_spawning_h, y = Percent_Motile, colour = Species),
    alpha = 0.25,
    size = 2,
    inherit.aes = FALSE
  ) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 1.2) +
  scale_colour_manual(
    values = c("D. pertusum" = "#f3c300", "D. dianthus" = "#9467bd"),
    labels = c(
      expression(italic("D. pertusum")),
      expression(italic("D. dianthus"))
    )
  ) +
  scale_fill_manual(
    values = c("D. pertusum" = "#f3c300", "D. dianthus" = "#9467bd"),
    labels = c(
      expression(italic("D. pertusum")),
      expression(italic("D. dianthus"))
    )
  ) +
  labs(
    x = "Time after spawning (hrs)",
    y = "Predicted Percent Motile (%)",
    colour = "Species",
    fill = "Species"
  ) +
  coord_cartesian(ylim = c(0, 35)) +
  theme_minimal(base_size = 12) +
  theme(
    axis.line = element_line(linewidth = 0.7, colour = "black"),
    axis.ticks = element_line(linewidth = 0.7),
    axis.ticks.length = unit(0.1, "cm")
  )

# -----------------------------
# Combined figure
# -----------------------------
fig_motility_time<-(p_motile_bayes_smooth + p_progressive_bayes_smooth) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "a", tag_prefix = "(", tag_suffix = ")")
fig_motility_time
head(db_progressive)
db_progressive %>%
  filter(time_after_spawning_h > 10 & time_after_spawning_h < 15)
```
check models fit
```{r}
# 1. Numerical convergence checks
summary(mod_bayes_progressive_smooth)
summary(mod_bayes_motile_smooth)

# 2. Trace plots / chain mixing
plot(mod_bayes_progressive_smooth)
plot(mod_bayes_motile_smooth)

# 3. Posterior predictive checks
pp_check(mod_bayes_progressive_smooth)
pp_check(mod_bayes_motile_smooth)

# Optional: a couple of extra PPC views, still consistent with manuscript style
pp_check(mod_bayes_progressive_smooth, type = "dens_overlay")
pp_check(mod_bayes_motile_smooth, type = "dens_overlay")
```

save plots

```{r}
ggsave(
  filename = "Fig_motility_time.tiff",
  plot = fig_motility_time,
  device = "tiff",
  width = 180,
  height = 90,
  units = "mm",
  dpi = 600,
  compression = "lzw",
  bg = "white"
)

# Optional vector version for your own editing / revisions
ggsave(
  filename = "Fig_motility_time.pdf",
  plot = fig_motility_time,
  width = 180,
  height = 90,
  units = "mm",
  bg = "white"
)
```

changing the 0 for NAs

```{r}
# Polynomial Bayesian model with male random intercept

library(dplyr)
library(ggplot2)
library(brms)

# Prepare raw data (convert 0 → NA for testing)
db2 <- db %>%
  mutate(
    Percent_Progressive = ifelse(
      Motile_Sperm == 0,
      NA,
      Percent_Progressive
    )) %>%
  filter(
    !is.na(Percent_Progressive),
    !is.na(time_after_spawning_h),
    !is.na(Species),
    !is.na(male)
  ) %>%
  mutate(
    prop_progressive = Percent_Progressive / 100,
    prop_progressive = pmin(pmax(prop_progressive, 0.001), 0.999),
    Species = factor(Species, levels = c("D. pertusum", "D. dianthus")),
    male = factor(male)
  )

# Priors
priors <- c(
  prior(normal(0, 1.5), class = "b"),
  prior(exponential(1), class = "sd"),
  prior(gamma(2, 0.1), class = "phi")
)

# Fit Bayesian polynomial model
mod_bayes_poly <- brm(
  formula = prop_progressive ~ Species * poly(time_after_spawning_h, 2) + (1 | male),
  data = db2,
  family = Beta(),
  prior = priors,
  chains = 4,
  cores = 4,
  iter = 4000,
  warmup = 2000,
  seed = 123,
  control = list(adapt_delta = 0.99, max_treedepth = 12)
)

# Model summary and posterior predictive check
print(summary(mod_bayes_poly))
pp_check(mod_bayes_poly)

# Prediction grid for population-level predictions
newdat_poly <- expand.grid(
  time_after_spawning_h = seq(
    min(db2$time_after_spawning_h),
    max(db2$time_after_spawning_h),
    length.out = 200
  ),
  Species = levels(db2$Species)
)

pred_poly <- fitted(mod_bayes_poly, newdata = newdat_poly, re_formula = NA)

plotdat_poly <- cbind(newdat_poly, as.data.frame(pred_poly)) %>%
  mutate(
    Estimate = Estimate * 100,
    Q2.5 = Q2.5 * 100,
    Q97.5 = Q97.5 * 100
  )

# Plot with raw points (note: now zeros are removed)
p_progressive_bayes_poly <- ggplot(
  plotdat_poly,
  aes(
    x = time_after_spawning_h,
    y = Estimate,
    colour = Species,
    fill = Species,
    group = Species
  )
) +
  geom_point(
    data = db2,
    aes(x = time_after_spawning_h, y = Percent_Progressive, colour = Species),
    alpha = 0.25,
    size = 2,
    inherit.aes = FALSE
  ) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 1.2) +
  scale_colour_manual(
    values = c("D. pertusum" = "#f3c300", "D. dianthus" = "#9467bd"),
    labels = c(expression(italic("D. pertusum")),
               expression(italic("D. dianthus")))
  ) +
  scale_fill_manual(
    values = c("D. pertusum" = "#f3c300", "D. dianthus" = "#9467bd"),
    labels = c(expression(italic("D. pertusum")),
               expression(italic("D. dianthus")))
  ) +
  labs(
    x = "Time after spawning (hrs)",
    y = "Predicted Percent Progressive (%)",
    colour = "Species",
    fill = "Species"
  ) +
  coord_cartesian(ylim = c(0, 30)) +
  theme_minimal(base_size = 12) +
  theme(
    axis.line = element_line(linewidth = 0.7, colour = "black"),
    axis.ticks = element_line(linewidth = 0.7),
    axis.ticks.length = unit(0.1, "cm")
  )

print(p_progressive_bayes_poly)
```
Trying with a thresshold of minimum 5 motile sperm measured to include them in the model

```{r}
# Polynomial Bayesian model with combined quality filters

library(dplyr)
library(ggplot2)
library(brms)

# Thresholds
min_total_threshold  <- 100
#min_motile_threshold <- 5
# Prepare filtered dataset
db2 <- db %>%
  filter(
    Total_Sperm >= min_total_threshold,
    #Motile_Sperm >= min_motile_threshold,
    !is.na(Percent_Progressive),
    !is.na(time_after_spawning_h),
    !is.na(Species),
    !is.na(male)
  ) %>%
  mutate(
    prop_progressive = Percent_Progressive / 100,
    prop_progressive = pmin(pmax(prop_progressive, 0.001), 0.999),
    Species = factor(Species, levels = c("D. pertusum", "D. dianthus")),
    male = factor(male)
  )

# Check what remains
print(nrow(db2))
print(table(db2$Species))
print(summary(db2$Total_Sperm))
print(summary(db2$Motile_Sperm))

# Priors
priors <- c(
  prior(normal(0, 1.5), class = "b"),
  prior(exponential(1), class = "sd"),
  prior(gamma(2, 0.1), class = "phi")
)

# Fit Bayesian polynomial model
mod_bayes_poly <- brm(
  prop_progressive ~ Species * poly(time_after_spawning_h, 2) + (1 | male),
  data = db2,
  family = Beta(),
  prior = priors,
  chains = 4,
  cores = 4,
  iter = 4000,
  warmup = 2000,
  seed = 123,
  control = list(adapt_delta = 0.99, max_treedepth = 12)
)

# Model checks
print(summary(mod_bayes_poly))
pp_check(mod_bayes_poly)

# Prediction grid
newdat_poly <- expand.grid(
  time_after_spawning_h = seq(
    min(db2$time_after_spawning_h),
    max(db2$time_after_spawning_h),
    length.out = 200
  ),
  Species = levels(db2$Species)
)

pred_poly <- fitted(mod_bayes_poly, newdata = newdat_poly, re_formula = NA)

plotdat_poly <- cbind(newdat_poly, as.data.frame(pred_poly)) %>%
  mutate(
    Estimate = Estimate * 100,
    Q2.5 = Q2.5 * 100,
    Q97.5 = Q97.5 * 100
  )

# Plot
p_progressive_bayes_poly <- ggplot(
  plotdat_poly,
  aes(
    x = time_after_spawning_h,
    y = Estimate,
    colour = Species,
    fill = Species,
    group = Species
  )
) +
  geom_point(
    data = db2,
    aes(x = time_after_spawning_h, y = Percent_Progressive, colour = Species),
    alpha = 0.25,
    size = 2,
    inherit.aes = FALSE
  ) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 1.2) +
  scale_colour_manual(
    values = c("D. pertusum" = "#f3c300", "D. dianthus" = "#9467bd"),
    labels = c(expression(italic("D. pertusum")),
               expression(italic("D. dianthus")))
  ) +
  scale_fill_manual(
    values = c("D. pertusum" = "#f3c300", "D. dianthus" = "#9467bd"),
    labels = c(expression(italic("D. pertusum")),
               expression(italic("D. dianthus")))
  ) +
  labs(
    x = "Time after spawning (hrs)",
    y = "Predicted Percent Progressive (%)",
    colour = "Species",
    fill = "Species"
  ) +
  coord_cartesian(ylim = c(0, 30)) +
  theme_minimal(base_size = 12) +
  theme(
    axis.line = element_line(linewidth = 0.7, colour = "black"),
    axis.ticks = element_line(linewidth = 0.7),
    axis.ticks.length = unit(0.1, "cm")
  )

print(p_progressive_bayes_poly)
```

averaging replicates

```{r}
# Aggregate replicate drops, filter low-quality observations,
# fit Bayesian polynomial model, and plot results

library(dplyr)
library(ggplot2)
library(brms)

# -----------------------------
# 1. Aggregate replicate drops
# -----------------------------
db_agg <- db %>%
  group_by(date, Species, tube, male, egg_present, time_after_spawning_h) %>%
  summarise(
    n_rows = n(),
    Total_Sperm = mean(Total_Sperm, na.rm = TRUE),
    Total_Paths = mean(Total_Paths, na.rm = TRUE),
    Slow_Sperm = mean(Slow_Sperm, na.rm = TRUE),
    Motile_Sperm = mean(Motile_Sperm, na.rm = TRUE),
    Progressive_Motile_Sperm = mean(Progressive_Motile_Sperm, na.rm = TRUE),
    Total_motile_sperm = mean(Total_motile_sperm, na.rm = TRUE),
    Percent_Motile = mean(Percent_Motile, na.rm = TRUE),
    Percent_Progressive = mean(Percent_Progressive, na.rm = TRUE),
    Percent_Total_Motile = mean(Percent_Total_Motile, na.rm = TRUE),
    .groups = "drop"
  )

# Optional quick checks
print(dim(db))
print(dim(db_agg))
print(head(db_agg))

# --------------------------------------
# 2. Apply thresholds after aggregation
# --------------------------------------
min_total_threshold  <- 100
min_motile_threshold <- 5

db2 <- db_agg %>%
  filter(
    #Total_Sperm >= min_total_threshold,
    #Motile_Sperm >= min_motile_threshold,
    !is.na(Percent_Progressive),
    !is.na(time_after_spawning_h),
    !is.na(Species),
    !is.na(male)
  ) %>%
  mutate(
    prop_progressive = Percent_Progressive / 100,
    prop_progressive = pmin(pmax(prop_progressive, 0.001), 0.999),
    Species = factor(Species, levels = c("D. pertusum", "D. dianthus")),
    male = factor(male)
  )

# Optional quick checks
print(nrow(db2))
print(table(db2$Species))
print(summary(db2$Total_Sperm))
print(summary(db2$Motile_Sperm))

# -----------------------------
# 3. Set priors
# -----------------------------
priors <- c(
  prior(normal(0, 1.5), class = "b"),
  prior(exponential(1), class = "sd"),
  prior(gamma(2, 0.1), class = "phi")
)

# -----------------------------
# 4. Fit Bayesian model
# -----------------------------
mod_bayes_poly <- brm(
  formula = prop_progressive ~ Species * poly(time_after_spawning_h, 2) + (1 | male),
  data = db2,
  family = Beta(),
  prior = priors,
  chains = 4,
  cores = 4,
  iter = 4000,
  warmup = 2000,
  seed = 123,
  control = list(adapt_delta = 0.99, max_treedepth = 12)
)

# -----------------------------
# 5. Model checks
# -----------------------------
print(summary(mod_bayes_poly))
pp_check(mod_bayes_poly)

# -----------------------------
# 6. Prediction grid
# -----------------------------
newdat_poly <- expand.grid(
  time_after_spawning_h = seq(
    min(db2$time_after_spawning_h),
    max(db2$time_after_spawning_h),
    length.out = 200
  ),
  Species = levels(db2$Species)
)

pred_poly <- fitted(mod_bayes_poly, newdata = newdat_poly, re_formula = NA)

plotdat_poly <- cbind(newdat_poly, as.data.frame(pred_poly)) %>%
  mutate(
    Estimate = Estimate * 100,
    Q2.5 = Q2.5 * 100,
    Q97.5 = Q97.5 * 100
  )

# -----------------------------
# 7. Plot
# -----------------------------
p_progressive_bayes_poly <- ggplot(
  plotdat_poly,
  aes(
    x = time_after_spawning_h,
    y = Estimate,
    colour = Species,
    fill = Species,
    group = Species
  )
) +
  geom_point(
    data = db2,
    aes(x = time_after_spawning_h, y = Percent_Progressive, colour = Species),
    alpha = 0.25,
    size = 2,
    inherit.aes = FALSE
  ) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 1.2) +
  scale_colour_manual(
    values = c("D. pertusum" = "#f3c300", "D. dianthus" = "#9467bd"),
    labels = c(expression(italic("D. pertusum")),
               expression(italic("D. dianthus")))
  ) +
  scale_fill_manual(
    values = c("D. pertusum" = "#f3c300", "D. dianthus" = "#9467bd"),
    labels = c(expression(italic("D. pertusum")),
               expression(italic("D. dianthus")))
  ) +
  labs(
    x = "Time after spawning (hrs)",
    y = "Predicted Percent Progressive (%)",
    colour = "Species",
    fill = "Species"
  ) +
  coord_cartesian(ylim = c(0, 30)) +
  theme_minimal(base_size = 12) +
  theme(
    axis.line = element_line(linewidth = 0.7, colour = "black"),
    axis.ticks = element_line(linewidth = 0.7),
    axis.ticks.length = unit(0.1, "cm")
  )

print(p_progressive_bayes_poly)
```


create a table for values of motility

```{r}

library(dplyr)

vars <- c(
  "Percent_Motile",
  "Percent_Progressive",
  "Percent_Total_Motile",
  "Avg_Motile_VAP", "Avg_Prog_Motile_VAP", "Avg_Tot_Motile_VAP",
  "percent_slow"
)

# Use db_binned, but DON'T replace NA with 0
species_bin_summary <- db_binned %>%
  filter(!is.na(time_bin)) %>%   # keep only rows that fell into a bin
  group_by(Species, time_bin) %>%
  summarise(
    across(
      all_of(vars),
      list(
        mean = ~mean(.x, na.rm = TRUE),
        sd   = ~sd(.x, na.rm = TRUE),
        n    = ~sum(!is.na(.x)),
        ci_l = ~{
          n <- sum(!is.na(.x)); m <- mean(.x, na.rm = TRUE); s <- sd(.x, na.rm = TRUE)
          if (n < 2) NA_real_ else m - qt(0.975, df = n-1) * s/sqrt(n)
        },
        ci_u = ~{
          n <- sum(!is.na(.x)); m <- mean(.x, na.rm = TRUE); s <- sd(.x, na.rm = TRUE)
          if (n < 2) NA_real_ else m + qt(0.975, df = n-1) * s/sqrt(n)
        }
      ),
      .names = "{.col}_{.fn}"
    ),
    .groups = "drop"
  ) %>%
  arrange(Species, time_bin)

species_bin_summary
```

editing so it looks nicer 
```{r}

species_bin_summary_table <- species_bin_summary %>%
  pivot_longer(
    cols = -c(Species, time_bin),
    names_to = c("metric", ".value"),
    names_pattern = "^(.*)_(mean|sd|n|ci_l|ci_u)$"
  ) %>%
  mutate(
    mean_ci = ifelse(
      is.na(ci_l),
      sprintf("%.2f", mean),
      sprintf("%.2f (%.2f–%.2f)", mean, ci_l, ci_u)
    )
  ) %>%
  select(Species, time_bin, metric, n, mean_ci) %>%
  arrange(metric, Species, time_bin)

```
reshape and format

```{r}


species_bin_summary_table <- species_bin_summary %>%
  pivot_longer(
    cols = -c(Species, time_bin),
    names_to = c("metric", ".value"),
    names_pattern = "^(.*)_(mean|sd|n|ci_l|ci_u)$"
  ) %>%
  filter(n>0) %>% 
  mutate(
    mean_ci = ifelse(
      is.na(ci_l),
      sprintf("%.2f", mean),
      sprintf("%.2f (%.2f–%.2f)", mean, ci_l, ci_u)
    ),
    sd_fmt = sprintf("%.2f", sd)
  ) %>%
  select(Species, time_bin, metric, n, sd_fmt, mean_ci) %>%
  rename(sd = sd_fmt) %>%
  arrange(Species, time_bin, metric)

```

metric names edit

```{r}
species_bin_summary_table <- species_bin_summary_table %>%
  mutate(
    metric = recode(
      metric,
      Percent_Motile          = "Percent motile (%)",
      Percent_Progressive     = "Percent progressive (%)",
      Percent_Total_Motile    = "Percent total motile (%)",
      Avg_Motile_VAP          = "Motile VAP (µm s⁻¹)",
      Avg_Prog_Motile_VAP     = "Progressive motile VAP (µm s⁻¹)",
      Avg_Tot_Motile_VAP      = "Total motile VAP (µm s⁻¹)",
      percent_slow            = "Percent slow (%)"
    )
  )

```

bins into columns

```{r}
species_bin_summary_table
```

order

```{r}
species_bin_summary_table <- species_bin_summary_table %>%
  mutate(
    time_bin = factor(
      time_bin,
      levels = c("0-5", "5-10", "10-15", "15-20",
                 "20-25", "25-30", "30-35", "35-40")
    )
  )

species_bin_summary_table <- species_bin_summary_table %>%
  arrange(time_bin, metric)
species_bin_summary_table
```
export

```{r}
# write.table(
#   species_bin_summary_table,
#   "tables/species_bin_summary_table.txt",
#   sep = "\t",
#   row.names = FALSE,
#   quote = FALSE
# )
```


Compact table for the paper

```{r}
library(dplyr)
library(tidyr)

# keep bin order consistent
bin_levels <- c("0-5","15-20","25-30")
db_binned <- db_binned %>%
  mutate(time_bin = factor(time_bin, levels = bin_levels))

tab_A <- db_binned %>%
  mutate(time_bin = factor(time_bin, levels = bin_levels)) %>%
  filter(!is.na(time_bin)) %>%   # <- prevents the NA row
  group_by(Species, time_bin) %>%
  summarise(
    motile_mean = mean(Percent_Motile, na.rm = TRUE),
    motile_sd   = sd(Percent_Motile, na.rm = TRUE),
    motile_n    = sum(!is.na(Percent_Motile)),
    motile_se   = motile_sd / sqrt(motile_n),
    motile_l    = motile_mean - qt(0.975, df = motile_n - 1) * motile_se,
    motile_u    = motile_mean + qt(0.975, df = motile_n - 1) * motile_se,

    prog_mean = mean(Percent_Progressive, na.rm = TRUE),
    prog_sd   = sd(Percent_Progressive, na.rm = TRUE),
    prog_n    = sum(!is.na(Percent_Progressive)),
    prog_se   = prog_sd / sqrt(prog_n),
    prog_l    = prog_mean - qt(0.975, df = prog_n - 1) * prog_se,
    prog_u    = prog_mean + qt(0.975, df = prog_n - 1) * prog_se,
    .groups = "drop"
  ) %>%
  mutate(
    Motile      = sprintf("%.2f (%.2f–%.2f)", motile_mean, motile_l, motile_u),
    Progressive = sprintf("%.2f (%.2f–%.2f)", prog_mean, prog_l, prog_u)
  ) %>%
  select(Species, time_bin, Motile, Progressive) %>%
  pivot_wider(
    names_from = Species,
    values_from = c(Motile, Progressive),
    names_glue = "{Species} {.value}"
  ) %>%
  arrange(time_bin) %>%
  mutate(across(-time_bin, ~replace_na(.x, "—"))) %>%
  rename(Time = time_bin)

tab_A


```

table B

```{r}
# ---- B) Velocity (µm s^-1) ----
tab_B <- db_binned %>%
  mutate(time_bin = factor(time_bin, levels = bin_levels)) %>%
  filter(!is.na(time_bin)) %>%   # <- prevents the NA row
  group_by(Species, time_bin) %>%
  summarise(
    # Motile VAP
    vap_mean = mean(Avg_Motile_VAP, na.rm = TRUE),
    vap_sd   = sd(Avg_Motile_VAP, na.rm = TRUE),
    vap_n    = sum(!is.na(Avg_Motile_VAP)),
    vap_se   = vap_sd / sqrt(vap_n),
    vap_l    = vap_mean - qt(0.975, df = vap_n - 1) * vap_se,
    vap_u    = vap_mean + qt(0.975, df = vap_n - 1) * vap_se,

    # Progressive VAP
    pvap_mean = mean(Avg_Prog_Motile_VAP, na.rm = TRUE),
    pvap_sd   = sd(Avg_Prog_Motile_VAP, na.rm = TRUE),
    pvap_n    = sum(!is.na(Avg_Prog_Motile_VAP)),
    pvap_se   = pvap_sd / sqrt(pvap_n),
    pvap_l    = pvap_mean - qt(0.975, df = pvap_n - 1) * pvap_se,
    pvap_u    = pvap_mean + qt(0.975, df = pvap_n - 1) * pvap_se,
    .groups = "drop"
  ) %>%
  mutate(
    `Motile VAP` = sprintf("%.2f (%.2f–%.2f)", vap_mean, vap_l, vap_u),
    `Prog. VAP`  = sprintf("%.2f (%.2f–%.2f)", pvap_mean, pvap_l, pvap_u)
  ) %>%
  select(Species, time_bin, `Motile VAP`, `Prog. VAP`) %>%
  pivot_wider(
    names_from = Species,
    values_from = c(`Motile VAP`, `Prog. VAP`),
    names_glue = "{Species} {.value}"
  ) %>%
  arrange(time_bin) %>%
  mutate(across(-time_bin, ~replace_na(.x, "—"))) %>%
  rename(Time = time_bin)

tab_B

```

export tables

```{r}
write.table(tab_A, "tables/TableX_A_motility_proportions.txt",
            sep = "\t", row.names = FALSE, quote = FALSE)

write.table(tab_B, "tables/TableX_B_velocity.txt",
            sep = "\t", row.names = FALSE, quote = FALSE)

```


