---
title: "Motility_paired_models_2026"
author: "Diego Moreno"
date: "2026-02-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
Load packages and Data
```{r}
library(dplyr)
library(stringr)
library(ggplot2)
library(brms)
library(ggdist)
library(tidybayes)
library(patchwork)
db<-read.delim("../Data/analysis_input_sperm_motility.txt")
head(db)
```

Build paired dataset for percentage responses

```{r}
egg_df <- db %>%
  filter(time_after_spawning_h < 15) %>%
  mutate(
    egg_present = factor(str_trim(tolower(egg_present)), levels = c("no", "yes")),
    pair_id = paste(Species, date, tube, male, time_after_spawning_h, sep = "_")
  )

valid_pairs <- egg_df %>%
  group_by(pair_id) %>%
  summarise(n_levels = n_distinct(egg_present), .groups = "drop") %>%
  filter(n_levels == 2) %>%
  pull(pair_id)

egg_df_paired <- egg_df %>%
  filter(pair_id %in% valid_pairs) %>%
  mutate(pair_id = factor(pair_id))

egg_df_paired %>% count(Species, egg_present)

```

Prepare percentage-response datasets
```{r}
prepare_prop_paired <- function(df, response_col) {
  df %>%
    filter(!is.na(.data[[response_col]])) %>%
    mutate(Proportion = .data[[response_col]] / 100) %>%
    group_by(Species) %>%
    mutate(
      N_species = n(),
      Proportion_t = (Proportion * (N_species - 1) + 0.5) / N_species
    ) %>%
    ungroup()
}

model_data_mot_prop  <- prepare_prop_paired(egg_df_paired, "Percent_Motile")
model_data_prog_prop <- prepare_prop_paired(egg_df_paired, "Percent_Progressive")
```


Define and fit paired beta models

```{r}
bf_prop <- bf(
  Proportion_t ~ egg_present + (1 | pair_id),
  phi ~ egg_present
)

priors_prop <- c(
  prior(normal(0, 1), class = "Intercept"),
  prior(normal(0, 0.5), class = "b"),
  prior(normal(2, 1), class = "Intercept", dpar = "phi"),
  prior(normal(0, 0.5), class = "b", dpar = "phi"),
  prior(student_t(3, 0, 2.5), class = "sd", group = "pair_id")
)

fit_prop_species <- function(df, seed) {
  brm(
    bf_prop,
    data = df,
    family = Beta(link = "logit"),
    prior = priors_prop,
    chains = 4, iter = 4000, warmup = 2000,
    control = list(adapt_delta = 0.99, max_treedepth = 13),
    seed = seed
  )
}

m_mot_dianthus <- fit_prop_species(
  filter(model_data_mot_prop, Species == "D. dianthus"), seed = 701)

m_mot_pertusum <- fit_prop_species(
  filter(model_data_mot_prop, Species == "D. pertusum"), seed = 702)

m_prog_dianthus <- fit_prop_species(
  filter(model_data_prog_prop, Species == "D. dianthus"), seed = 801)

m_prog_pertusum <- fit_prop_species(
  filter(model_data_prog_prop, Species == "D. pertusum"), seed = 802)
```
Diagnostics for percentage models

```{r}
eval_brms <- function(fit, label) {
  cat("\n============================\n", label, "\n============================\n")
  print(summary(fit))
  cat("Max Rhat:", max(rhat(fit), na.rm = TRUE), "\n")
  print(pp_check(fit))
}

eval_brms(m_mot_dianthus,  "Motile % — D. dianthus")
eval_brms(m_mot_pertusum,  "Motile % — D. pertusum")
eval_brms(m_prog_dianthus, "Progressive % — D. dianthus")
eval_brms(m_prog_pertusum, "Progressive % — D. pertusum")
```

Extract paired posterior differences for percentages

```{r}
extract_paired_diff_prop <- function(fit, species_label) {
  pair_levels <- unique(fit$data$pair_id)

  new_no <- data.frame(
    pair_id = pair_levels,
    egg_present = factor("no", levels = levels(fit$data$egg_present))
  )
  new_yes <- data.frame(
    pair_id = pair_levels,
    egg_present = factor("yes", levels = levels(fit$data$egg_present))
  )

  ep_no  <- posterior_epred(fit, newdata = new_no,  re_formula = NULL)
  ep_yes <- posterior_epred(fit, newdata = new_yes, re_formula = NULL)

  diff_draws <- rowMeans((ep_yes - ep_no) * 100, na.rm = TRUE)

  data.frame(Species = species_label, diff = diff_draws)
}

diff_mot <- bind_rows(
  extract_paired_diff_prop(m_mot_dianthus, "D. dianthus"),
  extract_paired_diff_prop(m_mot_pertusum, "D. pertusum")
)

diff_prog <- bind_rows(
  extract_paired_diff_prop(m_prog_dianthus, "D. dianthus"),
  extract_paired_diff_prop(m_prog_pertusum, "D. pertusum")
)
```

Build paired dataset for VAP responses

```{r}
motile_vap_min <- 20
motile_vap_max <- 79
prog_vap_min   <- 80

prepare_vap_paired <- function(df, response_col, lower = -Inf, upper = Inf) {
  tmp <- df %>%
    filter(
      time_after_spawning_h < 15,
      !is.na(.data[[response_col]]),
      .data[[response_col]] >= lower,
      .data[[response_col]] <= upper
    ) %>%
    mutate(
      egg_present = factor(str_trim(tolower(egg_present)), levels = c("no", "yes")),
      response = .data[[response_col]],
      event_id = paste(Species, drop_ID, date, tube, male, time_after_spawning_h, sep = "_")
    )

  valid_events <- tmp %>%
    group_by(event_id) %>%
    summarise(n_levels = n_distinct(egg_present), .groups = "drop") %>%
    filter(n_levels == 2) %>%
    pull(event_id)

  tmp %>%
    filter(event_id %in% valid_events) %>%
    mutate(
      event_id = factor(event_id),
      Species  = factor(Species, levels = c("D. dianthus", "D. pertusum"))
    )
}

model_data_mot_vap <- prepare_vap_paired(
  db, "Avg_Motile_VAP", lower = motile_vap_min, upper = motile_vap_max)

model_data_prog_vap <- prepare_vap_paired(
  db, "Avg_Prog_Motile_VAP", lower = prog_vap_min, upper = Inf)
```

Define and fit paired gamma models for VAP

```{r}
bf_vap <- bf(
  response ~ egg_present + (1 | event_id),
  shape ~ egg_present
)

priors_vap <- c(
  prior(normal(log(100), 1), class = "Intercept"),
  prior(normal(0, 0.3), class = "b"),
  prior(normal(3, 1), class = "Intercept", dpar = "shape"),
  prior(normal(0, 0.3), class = "b", dpar = "shape"),
  prior(student_t(3, 0, 2.5), class = "sd", group = "event_id")
)

fit_vap_species <- function(df, seed) {
  brm(
    bf_vap,
    data = df,
    family = Gamma(link = "log"),
    prior = priors_vap,
    chains = 4, iter = 4000, warmup = 2000,
    control = list(adapt_delta = 0.99, max_treedepth = 13),
    seed = seed
  )
}

m_mot_vap_dianthus <- fit_vap_species(
  filter(model_data_mot_vap, Species == "D. dianthus"), seed = 301)

m_mot_vap_pertusum <- fit_vap_species(
  filter(model_data_mot_vap, Species == "D. pertusum"), seed = 302)

m_prog_vap_dianthus <- fit_vap_species(
  filter(model_data_prog_vap, Species == "D. dianthus"), seed = 401)

m_prog_vap_pertusum <- fit_vap_species(
  filter(model_data_prog_vap, Species == "D. pertusum"), seed = 402)
```

Diagnostics for VAP models

```{r}
eval_brms(m_mot_vap_dianthus,  "Motile VAP — D. dianthus")
eval_brms(m_mot_vap_pertusum,  "Motile VAP — D. pertusum")
eval_brms(m_prog_vap_dianthus, "Progressive VAP — D. dianthus")
eval_brms(m_prog_vap_pertusum, "Progressive VAP — D. pertusum")
```
Extract paired posterior differences for VAP

```{r}
extract_paired_diff_vap <- function(fit, species_label) {
  event_levels <- unique(fit$data$event_id)

  new_no <- data.frame(
    event_id = event_levels,
    egg_present = factor("no", levels = levels(fit$data$egg_present))
  )
  new_yes <- data.frame(
    event_id = event_levels,
    egg_present = factor("yes", levels = levels(fit$data$egg_present))
  )

  mu_no  <- posterior_epred(fit, newdata = new_no,  re_formula = NULL)
  mu_yes <- posterior_epred(fit, newdata = new_yes, re_formula = NULL)

  diff_draws <- rowMeans(mu_yes - mu_no, na.rm = TRUE)

  data.frame(Species = species_label, diff = diff_draws)
}

diff_mot_vap <- bind_rows(
  extract_paired_diff_vap(m_mot_vap_dianthus, "D. dianthus"),
  extract_paired_diff_vap(m_mot_vap_pertusum, "D. pertusum")
)

diff_prog_vap <- bind_rows(
  extract_paired_diff_vap(m_prog_vap_dianthus, "D. dianthus"),
  extract_paired_diff_vap(m_prog_vap_pertusum, "D. pertusum")
)
```

functions for plotting

```{r}
species_colors <- c("D. dianthus" = "#9467bd", "D. pertusum" = "#f3c300")
species_levels <- c("D. dianthus", "D. pertusum")

summarise_diff <- function(df) {
  df %>%
    group_by(Species) %>%
    summarise(
      median  = median(diff, na.rm = TRUE),
      lower95 = ggdist::hdi(diff, .width = 0.95)[1],
      upper95 = ggdist::hdi(diff, .width = 0.95)[2],
      lower66 = ggdist::hdi(diff, .width = 0.66)[1],
      upper66 = ggdist::hdi(diff, .width = 0.66)[2],
      .groups = "drop"
    )
}

lab_triplet_diff <- function(sum_df, fmt = "%.1f", y_up = 0.14, y_med = -0.10) {
  list(
    geom_text(
      data = sum_df,
      aes(x = lower95, y = Species, label = sprintf(fmt, lower95)),
      inherit.aes = FALSE,
      size = 3.0, color = "grey30", hjust = 1.1,
      position = position_nudge(y = y_up)
    ),
    geom_text(
      data = sum_df,
      aes(x = upper95, y = Species, label = sprintf(fmt, upper95)),
      inherit.aes = FALSE,
      size = 3.0, color = "grey30", hjust = -0.1,
      position = position_nudge(y = y_up)
    ),
    geom_text(
      data = sum_df,
      aes(x = median, y = Species, label = sprintf(fmt, median)),
      inherit.aes = FALSE,
      size = 3.2, fontface = "bold", color = "black", vjust = 1.4,
      position = position_nudge(y = y_med)
    )
  )
}

plot_diffs_fancy <- function(draw_df, sum_df, xlab) {
  ggplot(draw_df, aes(x = diff, y = Species)) +
    stat_halfeye(
      aes(fill = Species, slab_alpha = after_stat(level)),
      .width = c(0.66, 0.95),
      point_interval = "median_hdi",
      trim = TRUE,
      interval_size = 0,
      point_size = 0
    ) +
    scale_slab_alpha_discrete(range = c(0.66, 0.95), guide = "none") +
    geom_segment(
      data = sum_df,
      aes(x = lower95, xend = upper95, y = Species, yend = Species),
      inherit.aes = FALSE,
      linewidth = 1,
      color = "black"
    ) +
    geom_segment(
      data = sum_df,
      aes(x = lower66, xend = upper66, y = Species, yend = Species),
      inherit.aes = FALSE,
      linewidth = 1.7,
      color = "black"
    ) +
    geom_point(
      data = sum_df,
      aes(x = median, y = Species),
      inherit.aes = FALSE,
      size = 2.4,
      color = "black"
    ) +
    geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.8, color = "grey40") +
    lab_triplet_diff(sum_df) +
    scale_fill_manual(values = species_colors, drop = FALSE) +
    labs(x = xlab, y = "Species") +
    theme_classic(base_size = 12) +
    theme(
      legend.position = "none",
      axis.text.y = element_text(face = "italic")
    ) +
    coord_cartesian(clip = "off")
}
```

build the panels


```{r}
diff_mot      <- diff_mot %>% mutate(Species = factor(Species, levels = species_levels))
diff_prog     <- diff_prog %>% mutate(Species = factor(Species, levels = species_levels))
diff_mot_vap  <- diff_mot_vap %>% mutate(Species = factor(Species, levels = species_levels))
diff_prog_vap <- diff_prog_vap %>% mutate(Species = factor(Species, levels = species_levels))

summary_mot      <- summarise_diff(diff_mot)
summary_prog     <- summarise_diff(diff_prog)
summary_mot_vap  <- summarise_diff(diff_mot_vap)
summary_prog_vap <- summarise_diff(diff_prog_vap)

plot_mot_prop_paired <- plot_diffs_fancy(
  diff_mot, summary_mot,
  "Difference (percentage points)"
)

plot_prog_prop_paired <- plot_diffs_fancy(
  diff_prog, summary_prog,
  "Difference (percentage points)"
)

plot_mot_vap_paired <- plot_diffs_fancy(
  diff_mot_vap, summary_mot_vap,
  expression("Difference in VAP ("*mu*"m s"^{-1}*")")
)

plot_prog_vap_paired <- plot_diffs_fancy(
  diff_prog_vap, summary_prog_vap,
  expression("Difference in VAP ("*mu*"m s"^{-1}*")")
)
```

Final 2x2 figure

```{r}
strip_titles <- function(p) {
  p + labs(title = NULL, subtitle = NULL, caption = NULL) +
    theme(
      plot.title = element_blank(),
      plot.subtitle = element_blank(),
      plot.caption = element_blank()
    )
}

theme_journal <- theme_minimal(base_size = 10, base_family = "Arial") +
  theme(
    legend.position = "none",
    legend.text = element_text(face = "italic"),
    strip.text = element_blank()
  )

pA <- strip_titles(plot_mot_prop_paired) +
  labs(x = "Difference (percentage points)", y = NULL) +
  theme_journal +
  theme(axis.text.y = element_text(face = "italic")) +
  scale_fill_manual(values = species_colors, drop = FALSE) +
  scale_color_manual(values = species_colors, drop = FALSE)

pB <- strip_titles(plot_prog_prop_paired) +
  labs(x = "Difference (percentage points)", y = NULL) +
  theme_journal +
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  ) +
  scale_fill_manual(values = species_colors, drop = FALSE) +
  scale_color_manual(values = species_colors, drop = FALSE)

pC <- strip_titles(plot_mot_vap_paired) +
  labs(x = expression("Difference in VAP ("*mu*"m s"^{-1}*")"), y = NULL) +
  theme_journal +
  theme(axis.text.y = element_text(face = "italic")) +
  scale_fill_manual(values = species_colors, drop = FALSE) +
  scale_color_manual(values = species_colors, drop = FALSE)

pD <- strip_titles(plot_prog_vap_paired) +
  labs(x = expression("Difference in VAP ("*mu*"m s"^{-1}*")"), y = NULL) +
  theme_journal +
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  ) +
  scale_fill_manual(values = species_colors, drop = FALSE) +
  scale_color_manual(values = species_colors, drop = FALSE)

fig_2x2 <- (pA | pB) / (pC | pD) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "a", tag_prefix = "(", tag_suffix = ")")

fig_2x2
```
export

```{r}

ggsave(
  filename = "../Figures/Fig_2x2_2.tiff",
  plot = fig_2x2,
  device = "tiff",
  width = 174,
  height = 120,
  units = "mm",
  dpi = 600,
  compression = "lzw",
  bg = "white"
)

quartz(
  type = "pdf",
  file = "../Figures/Fig_2x2_2.pdf",
  width = 174 / 25.4,
  height = 120 / 25.4
)
print(fig_2x2)
dev.off()
```

